---
title: "Complete API Reference with Examples"
author: "JoonHo Lee (jlee296@ua.edu)"
date: "`r Sys.Date()`"
description: >
  A comprehensive reference for all DPprior package functions, organized by 
  category with detailed signatures, parameters, return values, and usage examples.
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Complete API Reference with Examples}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
library(DPprior)
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5,
  fig.align = "center",
  out.width = "85%"
)
set.seed(42)
```

## Overview

This vignette provides a complete API reference for the DPprior package. All 
exported functions are documented with their full signatures, parameter 
descriptions, return values, and usage examples.

The package functions are organized into six categories:

| Category | Description | Key Functions |
|----------|-------------|---------------|
| **Core Elicitation** | Canonical target, K fit, and dual contracts | `DPprior_target_K()`, `DPprior_fit()`, `DPprior_dual_hard()`, `DPprior_dual_soft()` |
| **Exact Computation** | Stirling numbers and exact distributions | `compute_log_stirling()`, `pmf_K_given_alpha()` |
| **Approximation** | Closed-form A1 and Newton-based A2 methods | `DPprior_a1()`, `DPprior_a2_newton()` |
| **Weight Distribution** | First size-biased stick $W_{SB}$ and co-clustering $\rho$ | `mean_w1()`, `prob_wsb_exceeds()` |
| **Diagnostics** | Prior validation and estimand-specific weight assessment | `DPprior_diagnostics()`, `summary()` |
| **Utility** | Helper functions for conversions and integration | `vif_to_variance()`, `integrate_gamma()` |


---


## 1. Core Elicitation Functions

These are the primary user-facing functions for eliciting Gamma hyperpriors 
on the DP concentration parameter $\alpha$.

### 1.1 `DPprior_fit()`
**Unified Interface for Prior Elicitation**

The main entry point for eliciting a Gamma hyperprior. Automatically selects 
the appropriate algorithm based on input specification.

```r
DPprior_fit(
  J,                      # Sample size (required)
  mu_K = NULL,            # Target E[K_J], unless target_pmf is supplied
  var_K = NULL,           # Target Var(K_J) (optional)
  confidence = c("medium", "low", "high"),
  method = c("A2-MN", "A1", "A2-KL"),
  target_pmf = NULL,      # Optional custom PMF target for A2-KL
  check_diagnostics = TRUE,
  warn_dominance = NULL,  # Deprecated warning adapter
  M = 80L,                # Quadrature nodes
  verbose = FALSE,
  warning_policy = NULL,
  a1_projection = c("error", "nearest"),
  cv_K = NULL,
  K_interval = NULL,
  target_K = NULL,
  ...                     # Must be empty
)
```

**Parameters:**

| Parameter | Type | Description |
|-----------|------|-------------|
| `J` | Integer | Sample size (number of observations/sites). Must be $\geq 2$. |
| `mu_K` | Numeric | Target prior mean $\mathbb{E}[K_J]$. Moment workflows require $1 < \mu_K < J$; may be omitted when `target_pmf` is supplied. |
| `var_K` | Numeric | Target prior variance $\text{Var}(K_J)$. If `NULL`, computed from `confidence`. For fixed `mu_K`, must satisfy $\text{Var}(K_J) \leq (\mu_K - 1)(J - \mu_K)$. |
| `confidence` | Character | Qualitative uncertainty: `"low"` (VIF=5.0), `"medium"` (VIF=2.5), `"high"` (VIF=1.5), where $\text{Var}(K_J)=\text{VIF}(\mu_K - 1)$. |
| `method` | Character | Algorithm: `"A1"` (closed-form), `"A2-MN"` (Newton), `"A2-KL"` (KL minimization). |
| `target_pmf` | Numeric | Optional custom PMF target for `K_J`. Length may be `J` for support `1:J`, or `J+1` for support `0:J` when the `K = 0` entry is zero. If supplied and `method` is omitted, dispatches to `"A2-KL"`; explicit `mu_K` or `var_K` values must match the PMF-implied moments. |
| `check_diagnostics` | Logical | If `TRUE`, compute full diagnostics. |
| `warn_dominance` | Logical | Deprecated adapter. If `TRUE`, translates to the explicit policy $P(W_{SB}>0.5)>0.4$; it does not evaluate $W_{\max}$. Prefer `warning_policy`. |
| `M` | Integer | Number of Gauss-Laguerre quadrature nodes. |
| `verbose` | Logical | Print iteration progress. |
| `cv_K` | Numeric | Optional coefficient of variation route, `SD(K_J)/E(K_J)`. |
| `K_interval` | List | Optional canonical interval-target specification. |
| `target_K` | dpprior_K_target | Validated target from `DPprior_target_K()`; cannot be combined with scalar target arguments. |
| `...` | Dots | Must be empty. |

**Returns:** A canonical `dpprior.result/1` `DPprior_fit` object. Its
scientific and audit fields are:

| Component | Description |
|-----------|-------------|
| `parameters$a`, `parameters$b` | Gamma shape and rate |
| `J` | Sample size |
| `target$K$used`, `target$K$implied` | Authoritative requested/used target and implied target moments |
| `achieved$K` | Fresh achieved K evidence |
| `residuals` | Scientific residuals in their named units |
| `method` | Algorithm used |
| `status`, `usable`, `verified`, `message` | Decision/status quartet |
| `computation`, `verification`, `provenance` | Attempts, termination, independent checks, and audit trail |

**Examples:**

```{r dpprior-fit-examples}
# Primary workflow: construct, review, and fit one canonical target
target_K <- DPprior_target_K(J = 50, mu_K = 5, confidence = "medium")
fit1 <- DPprior_fit(
  J = 50, target_K = target_K, method = "A2-MN",
  check_diagnostics = FALSE
)
print(fit1)

# Scalar convenience route (the canonical target route above is preferred)
fit2 <- DPprior_fit(
  J = 50, mu_K = 5, var_K = 10, check_diagnostics = FALSE
)
cat("Gamma(", round(fit2$parameters$a, 4), ", ",
    round(fit2$parameters$b, 4), ")\n", sep = "")

# Using A1 closed-form (faster for large J)
target_A1 <- DPprior_target_K(J = 200, mu_K = 15, var_K = 30)
fit3 <- DPprior_fit(
  J = 200, target_K = target_A1, method = "A1",
  check_diagnostics = FALSE
)

# With diagnostics
target_diagnostics <- DPprior_target_K(J = 50, mu_K = 5, var_K = 8)
fit4 <- DPprior_fit(
  J = 50, target_K = target_diagnostics, check_diagnostics = TRUE
)

# Custom PMF target; omitted method dispatches to A2-KL
target_pmf <- dpois(1:30, lambda = 5)
target_pmf <- target_pmf / sum(target_pmf)
target_K_pmf <- DPprior_target_K(J = 30, target_pmf = target_pmf)
fit5 <- DPprior_fit(
  J = 30, target_K = target_K_pmf, check_diagnostics = FALSE
)
```


### 1.2 `DPprior_target_K()`
**Construct a Canonical Bounded-Discrete K Target**

```r
DPprior_target_K(
  J, mu_K = NULL, var_K = NULL, confidence = NULL, cv_K = NULL,
  K_interval = NULL, target_pmf = NULL, tolerance = 1e-9,
  root_control = NULL
)
```

Supply exactly one uncertainty source: `var_K`, `confidence`, `cv_K`,
`K_interval`, or `target_pmf`. The returned `dpprior.target/1` object retains
the exact support, original request, normalized and used values, implied
moments, tolerances, verification, and provenance. Pass it unchanged as
`target_K` to `DPprior_fit()`.

```{r dpprior-target-example}
target_K <- DPprior_target_K(J = 50, mu_K = 5, var_K = 8)
target_K$used[c("mu_K", "var_K")]
fit_K <- DPprior_fit(J = 50, target_K = target_K, method = "A2-MN")
```

### 1.3 `DPprior_dual_hard()`
**Verified Named Weight Inequality**

```r
DPprior_dual_hard(
  fit,
  constraint = list(
    metric = "wsb_tail", threshold = 0.5,
    relation = "<=", bound = 0.30
  ),
  constraint_tol = list(abs = 1e-6, rel = 1e-6),
  M = 80L, M_verify = NULL,
  log_bounds = c(-15, 15), control = list(),
  allow_approximate = FALSE, ...
)
```

Supported named metrics are `"wsb_tail"`, `"wsb_mean"`,
`"wsb_quantile"`, and the certified `"wmax_tail_upper"`. Hard mode has no
`lambda`. A decision-ready result requires its mode-specific contract,
`constraint$satisfied`, and an approved
`status`/`usable`/`verified`/`message` quartet.
Infeasible, unknown-feasibility, or unapproved approximate outcomes signal a
typed condition whose `result` retains the canonical evidence.

```{r dpprior-dual-hard-example}
fit_hard <- DPprior_dual_hard(
  fit = fit_K,
  constraint = list(
    metric = "wsb_tail", threshold = 0.5,
    relation = "<=", bound = 0.30
  )
)
fit_hard$constraint[c("satisfied", "residual", "tolerance")]
fit_hard[c("status", "usable", "verified", "message")]
```

### 1.4 `DPprior_dual_soft()`
**Fixed-Scale Soft Trade-off**

```r
DPprior_dual_soft(
  fit, target, lambda, max_iter = 100L,
  M_fit = 80L, M_verify = NULL,
  log_bounds = c(-15, 15), control = list(),
  allow_approximate = FALSE, start = NULL, ...
)
```

`target` names a weight metric, relation, and value. `lambda` is mandatory
and satisfies `0 < lambda <= 1`; it is a trade-off weight, not a constraint
probability or certificate. The `tradeoff` extension records the fixed
input-derived scales, component losses, and total objective. Soft objects do
not contain `constraint_satisfied`.

```{r dpprior-dual-soft-example}
fit_soft <- DPprior_dual_soft(
  fit = fit_K,
  target = list(
    metric = "wsb_tail", threshold = 0.5,
    relation = "target", value = 0.30
  ),
  lambda = 0.7
)
fit_soft$tradeoff[c("lambda", "K_loss", "weight_loss", "total_loss")]
fit_soft[c("status", "usable", "verified", "message")]
```

### 1.5 Legacy v2.x Compatibility: `DPprior_dual()`

`DPprior_dual()` remains available throughout v2.x for legacy equality-loss
workflows. It returns an explicitly approximate, unverified canonical result
and is not a hard-constraint certificate. It will not be removed before v3.0
and a migration review. New analyses should use the hard or soft API above.

```{r dpprior-dual-legacy-example, eval=FALSE}
legacy_fit <- DPprior_dual(
  fit = fit_K,
  w1_target = list(prob = list(threshold = 0.5, value = 0.30)),
  lambda = 0.7,
  loss_type = "adaptive"
)
legacy_fit[c("status", "usable", "verified", "message")]
```


---


## 2. Exact Computation Functions

These functions provide exact computation of Stirling numbers, PMFs, and 
moments using numerically stable log-space arithmetic.

### 2.1 Stirling Number Functions

#### `compute_log_stirling()`
**Pre-compute Log Stirling Numbers**

Computes the logarithm of unsigned Stirling numbers of the first kind 
$|s(J,k)|$ for all $J$ from 0 to `J_max` and $k$ from 0 to $J$.

```r
compute_log_stirling(J_max)
```

**Parameters:**

| Parameter | Type | Description |
|-----------|------|-------------|
| `J_max` | Integer | Maximum sample size. Must be $\leq 500$. |

**Returns:** A lower triangular matrix of dimension $(J_{max}+1) \times (J_{max}+1)$. 
Entry `[J+1, k+1]` contains $\log|s(J,k)|$ (using R's 1-based indexing).

**Examples:**

```{r stirling-examples}
# Pre-compute for J up to 100
logS <- compute_log_stirling(100)

# Access |s(10,3)| = 9450
s_10_3 <- exp(logS[11, 4])
cat("|s(10,3)| =", round(s_10_3), "\n")

# Verify row sum identity: sum_k |s(J,k)| = J!
J <- 6
row_sum <- sum(exp(logS[J+1, 2:(J+1)]))
cat("sum |s(6,k)| =", round(row_sum), ", 6! =", factorial(6), "\n")
```


### 2.2 Conditional PMF Functions

#### `pmf_K_given_alpha()`
**Antoniak Distribution PMF**

Computes the exact conditional PMF $P(K_J = k | \alpha)$ using the Antoniak 
distribution formula:
$$
P(K_J = k | \alpha) = \frac{|s(J,k)| \cdot \alpha^k}{(\alpha)_J}
$$

```r
pmf_K_given_alpha(J, alpha, logS, normalize = TRUE)
```

**Parameters:**

| Parameter | Type | Description |
|-----------|------|-------------|
| `J` | Integer | Sample size. |
| `alpha` | Numeric | Concentration parameter (scalar, $> 0$). |
| `logS` | Matrix | Pre-computed log-Stirling matrix. |
| `normalize` | Logical | If `TRUE`, ensure PMF sums to 1. |

**Returns:** Numeric vector of length $J+1$ containing $P(K_J = k | \alpha)$ 
for $k = 0, 1, \ldots, J$. Note: $P(K_J = 0 | \alpha) = 0$ always.

**Examples:**

```{r pmf-conditional-examples}
J <- 50
alpha <- 2.0
logS <- compute_log_stirling(J)

# Compute conditional PMF
pmf <- pmf_K_given_alpha(J, alpha, logS)

# Find mode
mode_k <- which.max(pmf) - 1
cat("Mode of K|α=2:", mode_k, "\n")

# Verify normalization
cat("PMF sum:", sum(pmf), "\n")
```


#### `cdf_K_given_alpha()`
**Conditional CDF**

```r
cdf_K_given_alpha(J, alpha, logS)
```

Returns the cumulative distribution function $F(k) = P(K_J \leq k | \alpha)$.


#### `quantile_K_given_alpha()`
**Conditional Quantile Function**

```r
quantile_K_given_alpha(p, J, alpha, logS)
```

Returns the smallest $k$ such that $P(K_J \leq k | \alpha) \geq p$.


### 2.3 Conditional Moment Functions

#### `mean_K_given_alpha()`
**Conditional Mean via Digamma**

Computes $\mathbb{E}[K_J | \alpha] = \alpha \{\psi(\alpha + J) - \psi(\alpha)\}$ 
using the digamma function.

```r
mean_K_given_alpha(J, alpha)
```

**Parameters:**

| Parameter | Type | Description |
|-----------|------|-------------|
| `J` | Integer | Sample size ($\geq 1$). |
| `alpha` | Numeric | Concentration parameter (vectorized). |

**Returns:** Numeric vector of conditional means.

**Examples:**

```{r mean-conditional-examples}
J <- 50

# Single alpha
mean_K_given_alpha(J, 2.0)

# Vectorized
alpha_seq <- c(0.5, 1, 2, 5, 10)
means <- mean_K_given_alpha(J, alpha_seq)
data.frame(alpha = alpha_seq, E_K = round(means, 2))
```


#### `var_K_given_alpha()`
**Conditional Variance via Trigamma**

Computes $\text{Var}(K_J | \alpha) = \mu_J(\alpha) - \alpha^2 \{\psi_1(\alpha) - \psi_1(\alpha + J)\}$.

```r
var_K_given_alpha(J, alpha)
```

**Key Property:** Conditional underdispersion always holds:
$$
0 < \text{Var}(K_J | \alpha) < \mathbb{E}[K_J | \alpha]
$$

**Examples:**

```{r var-conditional-examples}
J <- 50
alpha <- 2.0

mu <- mean_K_given_alpha(J, alpha)
sigma2 <- var_K_given_alpha(J, alpha)

cat("E[K|α=2] =", round(mu, 4), "\n")
cat("Var(K|α=2) =", round(sigma2, 4), "\n")
cat("Underdispersion ratio:", round(sigma2/mu, 4), "< 1 ✓\n")
```


#### `moments_K_given_alpha()`
**Both Moments in One Call**

```r
moments_K_given_alpha(J, alpha)
```

Returns a list with `mean` and `var`.


### 2.4 Marginal Distribution Functions

#### `exact_K_moments()`
**Marginal Moments under Gamma Hyperprior**

Computes $\mathbb{E}[K_J | a, b]$ and $\text{Var}(K_J | a, b)$ when 
$\alpha \sim \text{Gamma}(a, b)$ using Gauss-Laguerre quadrature.

```r
exact_K_moments(J, a, b, M = 80L)
```

**Parameters:**

| Parameter | Type | Description |
|-----------|------|-------------|
| `J` | Integer | Sample size. |
| `a` | Numeric | Gamma shape parameter. |
| `b` | Numeric | Gamma rate parameter. |
| `M` | Integer | Number of quadrature nodes. |

**Returns:** A list with components:

| Component | Description |
|-----------|-------------|
| `mean` | Marginal mean $M_1(a,b)$ |
| `var` | Marginal variance $V(a,b)$ |
| `sd` | Standard deviation |
| `cv` | Coefficient of variation |

**Examples:**

```{r exact-moments-examples}
# Compute marginal moments
result <- exact_K_moments(J = 50, a = 1.6, b = 1.22)
cat("E[K_50] =", round(result$mean, 4), "\n")
cat("Var(K_50) =", round(result$var, 4), "\n")
cat("CV(K_50) =", round(result$cv, 4), "\n")
```


#### `pmf_K_marginal()`
**Marginal PMF via Quadrature Mixture**

```r
pmf_K_marginal(J, a, b, logS, M = 80L)
```

Computes the marginal PMF by mixing conditional PMFs over $\alpha \sim \text{Gamma}(a,b)$:
$$
P(K_J = k | a, b) = \int_0^\infty P(K_J = k | \alpha) \cdot g_{a,b}(\alpha) \, d\alpha
$$

**Examples:**

```{r pmf-marginal-examples}
J <- 50
logS <- compute_log_stirling(J)

pmf_marginal <- pmf_K_marginal(J, a = 1.6, b = 1.22, logS)

# Find mode
mode_k <- which.max(pmf_marginal) - 1
cat("Mode of marginal K:", mode_k, "\n")
```


#### `summary_K_marginal()`
**Complete Marginal Distribution Summary**

```r
summary_K_marginal(J, a, b, logS, M = 80L, probs = c(0.05, 0.25, 0.5, 0.75, 0.95))
```

Returns mean, variance, mode, median, quantiles, PMF, and CDF.


---


## 3. Approximation Functions

### 3.1 `DPprior_a1()`
**A1 Closed-Form Approximation**

Fast closed-form solution using the negative binomial approximation to the 
marginal distribution of $K_J$.

```r
DPprior_a1(
  J,                      # Sample size
  mu_K,                   # Target mean
  var_K,                  # Target variance
  scaling = "log",        # "log", "harmonic", "digamma"
  epsilon = 1e-6          # Feasibility projection buffer
)
```

**Key Formulas:**
$$
a = \frac{(\mu_K - 1)^2}{\sigma^2_K - (\mu_K - 1)}, \quad
b = \frac{(\mu_K - 1) \cdot c_J}{\sigma^2_K - (\mu_K - 1)}
$$

where $c_J$ is the scaling constant (default: $\log J$).

**Feasibility Constraint:** Requires $\sigma^2_K > \mu_K - 1$.

**Examples:**

```{r a1-examples}
# Basic A1 fit
fit_a1 <- DPprior_a1(J = 50, mu_K = 5, var_K = 8)
cat("A1 solution: Gamma(", round(fit_a1$parameters$a, 4), ", ",
    round(fit_a1$parameters$b, 4), ")\n", sep = "")

# Compare scaling methods
for (s in c("log", "harmonic", "digamma")) {
  fit <- DPprior_a1(50, 5, 8, scaling = s)
  cat(sprintf("  %s: cJ = %.4f\n", s,
              fit$computation$scaling$values$cJ))
}
```


### 3.2 `DPprior_a2_newton()`
**A2-MN Newton Solver for Exact Moment Matching**

Finds $(a^*, b^*)$ such that the induced marginal moments exactly match targets.

```r
DPprior_a2_newton(
  J,                      # Sample size
  mu_K,                   # Target mean
  var_K,                  # Target variance
  a0 = NULL,              # Initial shape (from A1 if NULL)
  b0 = NULL,              # Initial rate (from A1 if NULL)
  tol_F = 1e-8,           # Residual tolerance
  tol_step = 1e-10,       # Step size tolerance
  max_iter = 20L,
  damping = TRUE,         # Backtracking line search
  use_fallback = TRUE,    # Nelder-Mead fallback
  M = 80L,
  verbose = FALSE
)
```

**Algorithm:**

1. Initialize from A1 closed-form
2. Log-parameterize: $\eta = (\log a, \log b)$
3. Newton iteration with score-based Jacobian
4. Backtracking line search for global convergence

**Convergence:** Typically achieves machine-precision accuracy in 3-8 iterations.

**Examples:**

```{r a2-newton-examples}
# Exact moment matching
fit <- DPprior_a2_newton(J = 50, mu_K = 5, var_K = 8, verbose = TRUE)

# Verify exact matching
achieved <- exact_K_moments(
  50, fit$parameters$a, fit$parameters$b
)
cat("\nTarget vs Achieved:\n")
cat("  E[K]:   5.000000 vs", sprintf("%.10f", achieved$mean), "\n")
cat("  Var(K): 8.000000 vs", sprintf("%.10f", achieved$var), "\n")
```


### 3.3 `DPprior_a2_kl()`
**A2-KL Distribution Matching via KL Divergence**

Minimizes Kullback-Leibler divergence between a target PMF and the induced 
marginal PMF of $K_J$.

```r
DPprior_a2_kl(
  J,                      # Sample size
  target,                 # Target PMF or list(mu_K, var_K)
  method = c("pmf", "chisq"),  # Target type
  max_iter = 100L,
  tol = 1e-6,
  M = 80L,
  verbose = FALSE
)
```

**Target Specification:**

| `method` | `target` Input | Description |
|----------|----------------|-------------|
| `"pmf"` | Numeric vector of length $J$ or $J+1$ | Direct PMF on $\{1, \ldots, J\}$, or on $\{0, \ldots, J\}$ with the $K=0$ entry required to be zero and then dropped |
| `"chisq"` | `list(mu_K = ..., var_K = ...)` | Discretized chi-square matching moments |

**Examples:**

```{r a2-kl-examples}
# Method 1: Moment-based target using chi-square discretization
fit_kl <- DPprior_a2_kl(J = 50, target = list(mu_K = 5, var_K = 8), 
                         method = "chisq")
cat("A2-KL solution: Gamma(", round(fit_kl$parameters$a, 4), ", ",
    round(fit_kl$parameters$b, 4), ")\n", sep = "")
cat("Final KL divergence:",
    format(fit_kl$residuals$distribution$kl, scientific = TRUE), "\n")

# Method 2: Direct PMF target
target_pmf <- discretize_chisq(J = 50, df = 6.25, scale = 0.8)
fit_kl2 <- DPprior_a2_kl(J = 50, target = target_pmf, method = "pmf")
```


---


## 4. Weight Distribution Functions

Functions for computing properties of the first size-biased stick $W_{SB}$
and the co-clustering probability $\rho = \sum_h w_h^2$.

### 4.1 First Size-Biased Weight $W_{SB}$ Functions

The historical function names use `w1`; scientifically, their estimand is the
first size-biased stick $W_{SB}$, not the largest realized weight $W_{\max}$.
It follows a compound distribution:
$$
W_{SB} | \alpha \sim \text{Beta}(1, \alpha), \quad \alpha \sim \text{Gamma}(a, b)
$$

#### `mean_w1()`
**Marginal Mean of $W_{SB}$**

```r
mean_w1(a, b, M = 80L)
```

Computes $\mathbb{E}[W_{SB} | a, b] = \mathbb{E}[1/(1+\alpha)]$ via quadrature.

**Examples:**

```{r mean-w1-example}
mean_w1(a = 1.6, b = 1.22)
```


#### `var_w1()`
**Marginal Variance of $W_{SB}$**

```r
var_w1(a, b, M = 80L)
```


#### `quantile_w1()`
**Marginal Quantiles of $W_{SB}$**

```r
quantile_w1(p, a, b, n_grid = 1000, M = 80L)
```

Computes quantiles via numerical inversion of the CDF.

**Examples:**

```{r quantile-w1-example}
# Median of the first size-biased weight
median_w1 <- quantile_w1(0.5, a = 1.6, b = 1.22)
cat("Median(W_SB) =", round(median_w1, 4), "\n")
```


#### `prob_wsb_exceeds()`
**Tail Probability $P(W_{SB} > x)$**

```r
prob_wsb_exceeds(threshold, a, b)
```

This is a tail probability for a size-biased cluster. It must not be reported as
the probability that the largest cluster exceeds the threshold. The historical
name `prob_w1_exceeds()` remains an explicit alias for the same $W_{SB}$
estimand. For $W_{\max}$, use `wmax_tail_bounds()` only when a certified bound
answers the scientific question.

**Examples:**

```{r prob-wsb-example}
# First-size-biased-weight tail probabilities
a <- 1.6; b <- 1.22
cat("P(W_SB > 0.5) =", round(prob_wsb_exceeds(0.5, a, b), 4), "\n")
cat("P(W_SB > 0.7) =", round(prob_wsb_exceeds(0.7, a, b), 4), "\n")
cat("P(W_SB > 0.9) =", round(prob_wsb_exceeds(0.9, a, b), 4), "\n")
```


#### `cdf_w1()` and `density_w1()`
**CDF and Density of $w_1$**

```r
cdf_w1(x, a, b, M = 80L)
density_w1(x, a, b, M = 80L)
```


### 4.2 Co-Clustering Probability $\rho$ Functions

The co-clustering probability $\rho = \sum_{h=1}^\infty w_h^2$ represents the 
probability that two randomly chosen observations belong to the same cluster.

#### `mean_rho()` and `var_rho()`
**Marginal Moments of $\rho$**

```r
mean_rho(a, b, M = 80L)
var_rho(a, b, M = 80L)
```

**Key Identity:** $\mathbb{E}[\rho | \alpha] = \mathbb{E}[W_{SB} | \alpha] = 1/(1+\alpha)$,
so $\mathbb{E}[\rho | a, b] = \mathbb{E}[W_{SB} | a, b]$. This equality of
means does not make $\rho$ interchangeable with either $W_{SB}$ or $W_{\max}$.

**Examples:**

```{r rho-examples}
a <- 1.6; b <- 1.22
cat("E[rho] =", round(mean_rho(a, b), 4), "\n")
cat("E[W_SB] =", round(mean_w1(a, b), 4), "(same mean as rho)\n")
cat("Var(rho) =", round(var_rho(a, b), 4), "\n")
cat("Var(W_SB) =", round(var_w1(a, b), 4), "(different!)\n")
```


#### `cv_rho()`
**Coefficient of Variation of $\rho$**

```r
cv_rho(a, b, M = 80L)
```


### 4.3 Conditional Weight Functions

#### `mean_rho_given_alpha()` and `var_rho_given_alpha()`

```r
mean_rho_given_alpha(alpha)   # Returns 1/(1 + alpha)
var_rho_given_alpha(alpha)    # Returns 2*alpha / ((1+alpha)^2 * (2+alpha) * (3+alpha))
```


---


## 5. Diagnostic Functions

### 5.1 `DPprior_diagnostics()`
**Comprehensive Prior Diagnostics**

Computes a full diagnostic report implementing the "unintended prior" checks
from Lee (2026, Section 4).

```r
DPprior_diagnostics(
  fit, thresholds = c(0.5, 0.9), warning_policy = NULL,
  M_verify = NULL, abs_tol = 1e-10, rel_tol = 1e-8,
  allow_approximate = FALSE
)
```

**Parameters:**

| Parameter | Type | Description |
|-----------|------|-------------|
| `fit` | DPprior_fit | Canonical `dpprior.result/1` fit; flat or legacy lists must first be upgraded explicitly. |
| `thresholds` | Numeric | Reserved summary-view thresholds; only `c(0.5, 0.9)` is accepted. Use `prob_wsb_exceeds()` for arbitrary $W_{SB}$ tails and `wmax_tail_bounds()` for certified $W_{max}$ bounds. |
| `warning_policy` | List | Optional exact ordered fields `estimand`, `direction`, `weight_threshold`, and `action_threshold`. |
| `M_verify` | Integer | Optional independent quadrature order. |
| `allow_approximate` | Logical | Explicitly retain an approximate bundle for review if a requested method contract cannot be verified; it never upgrades status or verification. |

**Returns:** A canonical `dpprior.result/1` object with exact class
`c("DPprior_diagnostics", "dpprior_result", "list")`, the 18 common fields,
and a `diagnostics` extension:

| Component | Description |
|-----------|-------------|
| `diagnostics$alpha` | Verified $\alpha$ mean and CV |
| `diagnostics$K` | Verified $K_J$ mean, variance, PMF, and quadrature order |
| `diagnostics$weights` | Verified first-size-biased-weight mean $E[W_{SB}]$ |
| `diagnostics$coclustering` | Verified $\rho$ mean and variance |
| `diagnostics$policy_results` | Exact $W_{SB}$ policy evidence or an indeterminate/backend-unavailable $W_{max}$ record |
| `diagnostics$warnings` | One warning per triggered named policy |

No qualitative risk category is inferred. `W_SB` and `W_max` are distinct
estimands. Canonical diagnostics make no direct $W_{max}$ point-tail claim;
an unavailable $W_{max}$ policy remains visibly indeterminate. The separate
`wmax_tail_upper` hard-dual route may use a certified upper bound.

**Examples:**

```{r diagnostics-example}
target_K <- DPprior_target_K(J = 50, mu_K = 5, var_K = 8)
fit <- DPprior_fit(
  J = 50, target_K = target_K, check_diagnostics = FALSE
)
diag <- DPprior_diagnostics(fit)
diag[c("status", "usable", "verified", "message")]
print(diag)
```


### 5.2 S3 Methods for `DPprior_fit`

#### `print.DPprior_fit()`

```r
print(fit)
```

Displays a concise summary of the elicitation result.


#### `summary.DPprior_fit()`

```r
summary(fit, print_output = TRUE)
```

**Returns:** An object of class `summary.DPprior_fit` with detailed statistics.


#### `plot.DPprior_fit()`

```r
plot(fit)
```

Creates a four-panel visualization dashboard:

- Panel A: Gamma prior density for $\alpha$
- Panel B: Marginal PMF of $K_J$
- Panel C: PDF of the first size-biased weight $W_{SB}$
- Panel D: Summary statistics table


### 5.3 Individual Plot Functions

```r
plot_alpha_prior(fit)   # Alpha distribution only
plot_K_prior(fit)       # K distribution only
plot_w1_prior(fit)      # First-size-biased-weight distribution
```


### 5.4 Weight Diagnostics and Deprecated Compatibility

#### `compute_weight_diagnostics()`

```r
weight_diag <- compute_weight_diagnostics(
  a, b, thresholds = c(0.5, 0.9), M = 80L
)
weight_diag$size_biased
weight_diag$maximum[c("status", "usable", "verified", "message")]
```

The `size_biased` component reports $W_{SB}$. The `maximum` component explicitly
records that no canonical direct $W_{\max}$ point-tail backend is available and
does not contain a substituted $W_{SB}$ value. Use `wmax_tail_bounds()` for the
separately named certified-upper-bound path. Numeric compatibility aliases refer
only to $W_{SB}$; no unqualified risk category is returned.

#### Deprecated compatibility: `check_dominance_risk()`

The ambiguous wrapper remains available for compatibility, emits a typed
deprecation warning, and evaluates only
$P(W_{SB}>\text{threshold})>\text{risk\_level}$. It does not evaluate
$W_{\max}$ or assign a category. New code should express the estimand and action
rule through `DPprior_diagnostics(warning_policy = ...)`.

```{r deprecated-check-dominance, eval=FALSE}
check_dominance_risk(a, b, threshold = 0.5, risk_level = 0.3)
```


---


## 6. Utility Functions

### 6.1 Variance Conversion Functions

#### `vif_to_variance()`
**Convert Variance Inflation Factor to Variance**

```r
vif_to_variance(mu_K, vif)
```

Computes $\sigma^2_K = \text{VIF} \times (\mu_K - 1)$, based on the marginal 
overdispersion relationship.

**Examples:**

```{r vif-example}
# VIF = 2 means variance is twice the Poisson-like baseline
var_K <- vif_to_variance(mu_K = 5, vif = 2)
cat("var_K =", var_K, "\n")
```


#### `confidence_to_vif()`
**Map Confidence Level to VIF**
  
```r
confidence_to_vif(confidence)
```

| Confidence | VIF | Interpretation |
|------------|-----|----------------|
| `"low"` | 5.0 | High uncertainty |
| `"medium"` | 2.5 | Moderate uncertainty |
| `"high"` | 1.5 | Low uncertainty |


#### `cv_alpha_to_variance()`
**Convert CV of $\alpha$ to Variance of $K$**

```r
cv_alpha_to_variance(mu_K, cv_alpha)
```


### 6.2 Scaling Functions

#### `compute_scaling_constant()`

```r
compute_scaling_constant(J, scaling = "log", mu_K = NULL)
```

Computes the scaling constant $c_J$ used in the A1 approximation:

| `scaling` | Formula |
|-----------|---------|
| `"log"` | $\log J$ |
| `"harmonic"` | $\sum_{i=1}^{J} 1/i$ |
| `"digamma"` | $\psi(J+1) + \gamma$ |


### 6.3 Quadrature Functions

#### `gauss_laguerre_nodes()`
**Compute Gauss-Laguerre Quadrature Nodes and Weights**

```r
gauss_laguerre_nodes(M, alpha_param = 0)
```

**Parameters:**

| Parameter | Type | Description |
|-----------|------|-------------|
| `M` | Integer | Number of quadrature nodes. |
| `alpha_param` | Numeric | Generalized Laguerre parameter (for Gamma(a,b), use $a-1$). |

**Returns:** List with `nodes`, `weights`, and `weights_log`.


#### `build_gamma_quadrature()`
**Build Quadrature for Gamma Distribution**

```r
build_gamma_quadrature(a, b, M = 80L)
```

Transforms standard Laguerre quadrature to integrate against Gamma(a,b).

**Returns:** List with `alpha_nodes` and `weights_normalized`.


#### `integrate_gamma()`
**Compute Expectation under Gamma Distribution**

```r
integrate_gamma(f, a, b, M = 80L)
```

Computes $\mathbb{E}_{\alpha \sim \text{Gamma}(a,b)}[f(\alpha)]$ using 
Gauss-Laguerre quadrature.

**Examples:**

```{r integrate-gamma-example}
# E[alpha] = a/b
a <- 2; b <- 0.5
E_alpha <- integrate_gamma(identity, a, b)
cat("E[alpha] via quadrature:", round(E_alpha, 6), "\n")
cat("E[alpha] exact (a/b):", a/b, "\n")
```


### 6.4 Log-Space Numerical Functions

#### `logsumexp()`
**Binary Log-Sum-Exp**

```r
logsumexp(a, b)
```

Computes $\log(\exp(a) + \exp(b))$ stably.


#### `logsumexp_vec()`
**Vectorized Log-Sum-Exp**

```r
logsumexp_vec(x)
```

Computes $\log(\sum_i \exp(x_i))$ for a vector.


### 6.5 KL Divergence Functions

#### `kl_divergence_pmf()`
**KL Divergence Between PMFs**

```r
kl_divergence_pmf(p, q, eps = 1e-15)
```


#### `kl_divergence_K()`
**KL Divergence for $K_J$ Distributions**

```r
kl_divergence_K(target_pmf, a, b, J, M = 80L)
```


#### `discretize_chisq()`
**Create Target PMF from Chi-Square**

```r
discretize_chisq(J, df, scale = 1)
```

Discretizes a (scaled) chi-square distribution onto $\{1, \ldots, J\}$.


---


## 7. Verification Functions

The package includes extensive verification functions for development and 
testing. These are typically not needed by end users but are available for 
those who wish to validate computations.

### 7.1 Module Verification Functions

| Function | Description |
|----------|-------------|
| `validate_stirling(logS)` | Verify Stirling numbers against known values |
| `verify_stirling_row_sum(logS)` | Verify $\sum_k |s(J,k)| = J!$ |
| `verify_underdispersion(J, alpha_values)` | Verify $\text{Var} < \text{Mean}$ |
| `verify_pmf_moments(J, alpha, logS)` | Verify PMF matches closed-form moments |
| `verify_marginal_moments(J, a, b)` | Verify marginal moment properties |
| `verify_a1_roundtrip(fit)` | Verify A1 parameters recover targets |
| `verify_a2_moment_matching(J, mu_K, var_K)` | Verify A2 achieves exact matching |

**Examples:**

```{r verification-examples}
# Verify Stirling numbers
logS <- compute_log_stirling(20)
validate_stirling(logS, verbose = TRUE)

# Verify underdispersion
verify_underdispersion(50, alpha_values = c(0.5, 1, 2, 5))
```


---


## 8. Function Quick Reference

### By Task

**"I want to elicit a prior based on expected clusters"**
```r
target_K <- DPprior_target_K(J = 50, mu_K = 5, confidence = "medium")
fit <- DPprior_fit(J = 50, target_K = target_K, method = "A2-MN")
```

**"I need a verified hard inequality for $W_{SB}$"**
```r
fit <- DPprior_dual_hard(
  fit,
  constraint = list(
    metric = "wsb_tail", threshold = 0.5,
    relation = "<=", bound = 0.3
  )
)
```

**"I want an explicit K/weight trade-off"**
```r
fit <- DPprior_dual_soft(
  fit,
  target = list(
    metric = "wsb_tail", threshold = 0.5,
    relation = "target", value = 0.3
  ),
  lambda = 0.7
)
```

**"I want to assess a named weight estimand"**
```r
prob_wsb_exceeds(0.5, fit$parameters$a, fit$parameters$b)
diag <- DPprior_diagnostics(fit)
diag[c("status", "usable", "verified", "message")]
```

**"I want to compute the exact PMF of K"**
```r
logS <- compute_log_stirling(J)
pmf <- pmf_K_marginal(
  J, fit$parameters$a, fit$parameters$b, logS
)
```

**"I want to visualize my prior"**
```r
plot(fit)  # Full dashboard
plot_K_prior(fit)  # K distribution only
```


### Alphabetical Index

| Function | Category | Description |
|----------|----------|-------------|
| `build_gamma_quadrature()` | Utility | Quadrature for Gamma |
| `cdf_K_given_alpha()` | Exact | Conditional CDF |
| `cdf_w1()` | Weights | CDF of $w_1$ |
| `check_dominance_risk()` | Deprecated compatibility | Ambiguous-name wrapper for an explicit $W_{SB}$ policy |
| `compute_log_stirling()` | Exact | Stirling numbers |
| `compute_scaling_constant()` | Utility | Scaling $c_J$ |
| `confidence_to_vif()` | Utility | Confidence to VIF |
| `cv_alpha_to_variance()` | Utility | CV to variance |
| `cv_rho()` | Weights | CV of $\rho$ |
| `discretize_chisq()` | Utility | Chi-square to PMF |
| `DPprior_a1()` | Approximation | A1 closed-form |
| `DPprior_a2_kl()` | Approximation | A2-KL solver |
| `DPprior_a2_newton()` | Approximation | A2-MN Newton |
| `DPprior_diagnostics()` | Diagnostics | Full diagnostics |
| `DPprior_dual_hard()` | Core | Verified hard inequality |
| `DPprior_dual_soft()` | Core | Fixed-scale soft trade-off |
| `DPprior_fit()` | Core | Main entry point |
| `DPprior_target_K()` | Core | Canonical K target constructor |
| `exact_K_moments()` | Exact | Marginal moments |
| `gauss_laguerre_nodes()` | Utility | Quadrature nodes |
| `integrate_gamma()` | Utility | Expectation under Gamma |
| `kl_divergence_K()` | Utility | KL for K |
| `kl_divergence_pmf()` | Utility | KL between PMFs |
| `logsumexp()` | Utility | Binary log-sum-exp |
| `logsumexp_vec()` | Utility | Vector log-sum-exp |
| `mean_K_given_alpha()` | Exact | Conditional mean |
| `mean_rho()` | Weights | Marginal $\mathbb{E}[\rho]$ |
| `mean_rho_given_alpha()` | Weights | Conditional $\mathbb{E}[\rho|\alpha]$ |
| `mean_w1()` | Weights | Marginal $\mathbb{E}[w_1]$ |
| `moments_K_given_alpha()` | Exact | Both conditional moments |
| `density_w1()` | Weights | Density of $w_1$ |
| `plot.DPprior_fit()` | Diagnostics | Dashboard plot |
| `plot_alpha_prior()` | Diagnostics | Alpha plot |
| `plot_K_prior()` | Diagnostics | K plot |
| `plot_w1_prior()` | Diagnostics | $w_1$ plot |
| `pmf_K_given_alpha()` | Exact | Conditional PMF |
| `pmf_K_marginal()` | Exact | Marginal PMF |
| `print.DPprior_fit()` | Diagnostics | Print method |
| `prob_w1_exceeds()` | Weights | Historical alias for the $W_{SB}$ tail |
| `prob_wsb_exceeds()` | Weights | Explicitly named $W_{SB}$ tail probability |
| `quantile_K_given_alpha()` | Exact | Conditional quantile |
| `quantile_w1()` | Weights | Quantiles of $w_1$ |
| `summary.DPprior_fit()` | Diagnostics | Summary method |
| `summary_K_marginal()` | Exact | Full K summary |
| `var_K_given_alpha()` | Exact | Conditional variance |
| `var_rho()` | Weights | Marginal $\text{Var}(\rho)$ |
| `var_rho_given_alpha()` | Weights | Conditional $\text{Var}(\rho|\alpha)$ |
| `var_w1()` | Weights | Marginal $\text{Var}(w_1)$ |
| `vif_to_variance()` | Utility | VIF to variance |


---


## References

Antoniak, C. E. (1974). Mixtures of Dirichlet processes with applications to 
Bayesian nonparametric problems. *The Annals of Statistics*, 2(6), 1152-1174.

Golub, G. H., & Welsch, J. H. (1969). Calculation of Gauss quadrature rules. 
*Mathematics of Computation*, 23(106), 221-230.

Lee, J. (2026). Design-conditional prior elicitation for Dirichlet process
mixtures. *arXiv preprint* arXiv:2602.06301.

Lee, J., Che, J., Rabe-Hesketh, S., Feller, A., & Miratrix, L. (2025). 
Improving the estimation of site-specific effects and their distribution in 
multisite trials. *Journal of Educational and Behavioral Statistics*, 50(5), 
731-764.


---


*For additional documentation, see the [package vignettes](index.html) or 
visit the [GitHub repository](https://github.com/joonho112/DPprior).*
