---
title: "Case Studies: Multisite Trials and Meta-Analysis"
author: "JoonHo Lee (jlee296@ua.edu)"
date: "`r Sys.Date()`"
description: >
  Real-world applications of DPprior in educational multisite trials, 
  medical meta-analysis, and large-scale policy evaluation.
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Case Studies: Multisite Trials and Meta-Analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
library(DPprior)
library(ggplot2)
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5,
  fig.align = "center",
  out.width = "85%"
)
set.seed(42)

# Define a consistent color palette
palette_2 <- c("#E41A1C", "#377EB8")
palette_3 <- c("#E41A1C", "#377EB8", "#4DAF4A")
palette_4 <- c("#E41A1C", "#377EB8", "#4DAF4A", "#984EA3")
```

## Overview

This vignette presents three case studies demonstrating how to apply 
the DPprior package in real-world research contexts. Each case study 
walks through the complete workflow---from substantive considerations 
to final prior specification---using published research as a foundation.

The three case studies span different domains and data structures:

| Case Study | Domain | J (Sites/Studies) | Key Characteristic |
|------------|--------|-------------------|--------------------|
| **1. Conditional Cash Transfer Trial** | Education/Policy | 38 | Moderate sites, dual-anchor refinement |
| **2. Brief Alcohol Interventions** | Public Health | 117 | Large meta-analysis, multiple heterogeneity sources |
| **3. Alabama Pre-K Value-Added** | Education Policy | 500 (of 847) | Large-scale policy evaluation |

By working through these examples, you will learn how to:

1. Translate substantive domain knowledge into cluster count expectations
2. Choose appropriate calibration methods for different settings
3. Apply diagnostics to verify prior behavior
4. Conduct sensitivity analysis across specifications
5. Report prior elicitation transparently in publications


## 1. Case Study: Conditional Cash Transfer Multisite Trial

### 1.1 Research Context

Our first case study draws from the multisite conditional cash transfer (CCT) 
experiment analyzed in [Lee et al. (2025)](https://doi.org/10.3102/10769986241254286). 
The Conditional Subsidies for School Attendance program, implemented in 
Bogotá, Colombia ([Barrera-Osorio et al., 2019](https://doi.org/10.1257/app.20170008)), 
conducted randomized experiments across 38 sites in the San Cristobal district.

**Study characteristics:**

- **Sample size**: 6,506 participants nested within 38 sites
- **Average site size**: 171.2 participants
- **Coefficient of variation in site sizes**: 0.67 (range: 23 to 484)
- **Outcomes**: Secondary school enrollment and on-time graduation
- **Key feature**: Large within-site information, limited between-site variation

### 1.2 The Inferential Goals

The researchers sought to understand the distribution of site-specific 
treatment effects $\tau_j$, with particular interest in:

1. **Estimating the variance** of the prior distribution $G$
2. **Identifying extreme sites** (10th and 90th percentiles)
3. **Understanding heterogeneity** patterns across sites

Given the estimated cross-site effect standard deviation of $\sigma = 0.04$ 
to $0.06$ (in effect size units) and relatively high within-site precision, 
this setting exhibits **small between-site information and large within-site 
information**---a configuration where prior choice substantially influences 
posterior inference.

### 1.3 Initial Prior Elicitation

**Step 1: Determining the expected number of clusters**

When eliciting prior beliefs about cluster structure, researchers might 
consider several substantive factors:

- The intervention was implemented uniformly across sites, suggesting limited 
  structural heterogeneity
- However, local context (school culture, neighborhood characteristics) might 
  create differential responses
- Previous multisite education trials typically exhibit 3--8 distinct response 
  patterns

Based on these considerations, a reasonable expectation is approximately 
$\mu_K = 5$ clusters among the 38 sites.

```{r cct-step1}
# Define the study context
J_cct <- 38
mu_K_cct <- 5

cat("Study context:\n")
cat("  Number of sites (J):", J_cct, "\n")
cat("  Expected clusters (μ_K):", mu_K_cct, "\n")
cat("  Ratio (μ_K/J):", round(mu_K_cct / J_cct, 3), "\n")
```

**Step 2: Expressing uncertainty**

Given the novelty of this specific intervention context and the moderate 
sample of sites, we should express substantial uncertainty. A "low" 
confidence level is appropriate:

```{r cct-step2}
# Initial fit with a reviewable low-confidence target
target_cct <- DPprior_target_K(
  J = J_cct, mu_K = mu_K_cct, confidence = "low"
)
fit_cct <- DPprior_fit(
  J = J_cct,
  target_K = target_cct,
  method = "A2-MN",
  M = 160L,
  check_diagnostics = FALSE
)

print(fit_cct)
```

### 1.4 Comprehensive Diagnostics

Before accepting this prior, we need to verify its behavior across multiple 
dimensions. The `DPprior_diagnostics()` function provides a comprehensive 
assessment:

```{r cct-diagnostics}
# This low-confidence case can require an explicitly reviewable approximate
# diagnostic bundle; its status quartet is reported below rather than upgraded.
diag_cct <- DPprior_diagnostics(fit_cct, allow_approximate = TRUE)
alpha_cct <- diag_cct$diagnostics$alpha
K_cct <- diag_cct$diagnostics$K
wsb_cct <- diag_cct$diagnostics$weights
rho_cct <- diag_cct$diagnostics$coclustering
alpha_sd_cct <- sqrt(fit_cct$parameters$a) / fit_cct$parameters$b
alpha_q_cct <- qgamma(
  c(q5 = 0.05, q50 = 0.50, q95 = 0.95),
  shape = fit_cct$parameters$a, rate = fit_cct$parameters$b
)
K_mode_cct <- which.max(K_cct$pmf)
wsb_median_cct <- quantile_w1(
  0.5, fit_cct$parameters$a, fit_cct$parameters$b
)
wsb_tail_cct <- c(
  threshold_0.5 = prob_wsb_exceeds(
    0.5, fit_cct$parameters$a, fit_cct$parameters$b
  ),
  threshold_0.9 = prob_wsb_exceeds(
    0.9, fit_cct$parameters$a, fit_cct$parameters$b
  )
)

cat("=== Comprehensive Diagnostic Report ===\n\n")

cat("1. ALPHA DISTRIBUTION\n")
cat("   Mean:     ", round(alpha_cct$mean, 3), "\n")
cat("   SD:       ", round(alpha_sd_cct, 3), "\n")
cat("   CV:       ", round(alpha_cct$CV, 3), "\n")
cat("   5th %ile: ", round(alpha_q_cct["q5"], 3), "\n")
cat("   50th %ile:", round(alpha_q_cct["q50"], 3), "\n")
cat("   95th %ile:", round(alpha_q_cct["q95"], 3), "\n")

cat("\n2. CLUSTER COUNT (K)\n")
cat("   E[K]:     ", round(K_cct$mean, 3), "\n")
cat("   Var(K):   ", round(K_cct$variance, 3), "\n")
cat("   SD(K):    ", round(sqrt(K_cct$variance), 3), "\n")
cat("   Mode(K):  ", K_mode_cct, "\n")

cat("\n3. WEIGHT BEHAVIOR\n")
cat("   E[W_SB]:        ", round(wsb_cct$mean, 3), "\n")
cat("   Median(W_SB):   ", round(wsb_median_cct, 3), "\n")
cat("   P(W_SB > 0.5):  ",
    round(wsb_tail_cct["threshold_0.5"], 3), "\n")
cat("   P(W_SB > 0.9):  ",
    round(wsb_tail_cct["threshold_0.9"], 3), "\n")
cat("   W_SB status:    ", wsb_cct$status, "\n")

cat("\n4. CO-CLUSTERING PROBABILITY (ρ)\n")
cat("   E[ρ]:         ", round(rho_cct$mean, 3), "\n")
cat("   Interpretation: Two random sites have a", 
    round(100 * rho_cct$mean), "% chance of sharing a cluster\n")
cat("\nBundle status:\n")
print(diag_cct[c("status", "usable", "verified", "message")])
```

### 1.5 Visualizing the Initial Prior

The DPprior package provides several visualization functions to help 
understand prior behavior. Let's examine each component:

**Alpha distribution:**

```{r cct-viz-alpha, fig.height=4, fig.cap="Prior distribution on the concentration parameter α for the CCT study."}
plot_alpha_prior(fit_cct)
```

**Cluster count distribution:**

```{r cct-viz-K, fig.height=4, fig.cap="Prior PMF of the number of clusters K for the CCT study (J = 38)."}
plot_K_prior(fit_cct)
```

**First size-biased weight distribution:**

```{r cct-viz-w1, fig.height=4, fig.cap="Prior distribution of the first size-biased weight W_SB. The dashed line marks the reviewed mass threshold 0.5."}
plot_w1_prior(fit_cct)
```

**Complete dashboard:**

```{r cct-viz-dashboard, fig.height=8, fig.cap="Complete prior elicitation dashboard for the CCT multisite trial (J = 38, μ_K = 5)."}
plot(fit_cct)
```

### 1.6 Addressing a Large $W_{SB}$ Tail Probability

The diagnostics reveal an important concern: the named tail probability
$P(W_{SB}>0.5)=`r round(wsb_tail_cct["threshold_0.5"], 2)`$.
This means there is a `r round(100 * wsb_tail_cct["threshold_0.5"])`%
prior probability that the size-biased first stick-breaking cluster mass
exceeds one half.

In a setting with limited between-site information, this large
$P(W_{SB}>0.5)$ can lead to overshrinkage in posterior inference. The
dual-anchor framework allows us to constrain that named tail probability while
maintaining our beliefs about the expected number of clusters.

**Applying dual-anchor refinement:**

We require $P(W_{SB} > 0.5) \leq 0.25$ while keeping
$K$ as the optimization objective while independently certifying the weight
inequality:

```{r cct-dual-anchor}
# Apply a verified hard bound to the named W_SB tail probability
fit_cct_dual <- DPprior_dual_hard(
  fit = fit_cct,
  constraint = list(
    metric = "wsb_tail", threshold = 0.5,
    relation = "<=", bound = 0.25
  )
)

cat("Original prior:    Gamma(a =", round(fit_cct$parameters$a, 4),
    ", b =", round(fit_cct$parameters$b, 4), ")\n")
cat("Dual-anchor prior: Gamma(a =", round(fit_cct_dual$parameters$a, 4),
    ", b =", round(fit_cct_dual$parameters$b, 4), ")\n")
cat("Constraint verified:", fit_cct_dual$constraint$satisfied, "\n")
fit_cct_dual[c("status", "usable", "verified", "message")]
```

**Verifying the improvement:**

```{r cct-dual-diagnostics}
# The hard-fit object already carries its verified K and named weight evidence.
K_cct_dual <- fit_cct_dual$achieved$K
alpha_mean_cct_dual <- with(
  fit_cct_dual$parameters, a / b
)
wsb_mean_cct_dual <- mean_w1(
  fit_cct_dual$parameters$a, fit_cct_dual$parameters$b
)
wsb_tail_cct_dual <- fit_cct_dual$achieved$weight$value

# Create comparison table
comparison_df <- data.frame(
  Metric = c(
    "E[α]", "E[K]", "Var(K)", "E[W_SB]", "P(W_SB > 0.5)",
    "Diagnostic status"
  ),
  Original = c(
    round(alpha_cct$mean, 3),
    round(K_cct$mean, 2),
    round(K_cct$variance, 2),
    round(wsb_cct$mean, 3),
    round(wsb_tail_cct["threshold_0.5"], 3),
    diag_cct$status
  ),
  Dual_Anchor = c(
    round(alpha_mean_cct_dual, 3),
    round(K_cct_dual$mean, 2),
    round(K_cct_dual$variance, 2),
    round(wsb_mean_cct_dual, 3),
    round(wsb_tail_cct_dual, 3),
    fit_cct_dual$status
  )
)

knitr::kable(
  comparison_df,
  col.names = c("Metric", "K-only Prior", "Dual-Anchor Prior"),
  caption = "Comparison of K-only and dual-anchor priors for the CCT study"
)
```

### 1.7 Comparing K-only vs Dual-Anchor Priors

Let's visualize the difference between the two priors:

```{r cct-comparison-viz, fig.height=6, fig.cap="Comparison of K-only and dual-anchor priors: the dual-anchor approach reduces P(W_SB > 0.5) while maintaining similar cluster count expectations."}
# Prepare data for K PMF comparison
logS_cct <- compute_log_stirling(J_cct)

pmf_original <- pmf_K_marginal(
  J_cct, fit_cct$parameters$a, fit_cct$parameters$b, logS = logS_cct
)
pmf_dual <- pmf_K_marginal(
  J_cct, fit_cct_dual$parameters$a, fit_cct_dual$parameters$b,
  logS = logS_cct
)

k_compare_df <- data.frame(
  K = rep(seq_along(pmf_original), 2),
  Probability = c(pmf_original, pmf_dual),
  Prior = rep(c("K-only", "Dual-anchor"), each = length(pmf_original))
)
k_compare_df$Prior <- factor(k_compare_df$Prior, levels = c("K-only", "Dual-anchor"))

# Prepare data for w1 comparison
w1_grid <- seq(0, 1, length.out = 200)
w1_dens_original <- sapply(w1_grid, function(w) {
  density_w1(w, fit_cct$parameters$a, fit_cct$parameters$b)
})
w1_dens_dual <- sapply(w1_grid, function(w) {
  density_w1(w, fit_cct_dual$parameters$a, fit_cct_dual$parameters$b)
})

w1_compare_df <- data.frame(
  w1 = rep(w1_grid, 2),
  Density = c(w1_dens_original, w1_dens_dual),
  Prior = rep(c("K-only", "Dual-anchor"), each = length(w1_grid))
)
w1_compare_df$Prior <- factor(w1_compare_df$Prior, levels = c("K-only", "Dual-anchor"))

# Create plots
p1 <- ggplot(k_compare_df[k_compare_df$K <= 15, ], 
             aes(x = K, y = Probability, fill = Prior)) +
  geom_col(position = "dodge", alpha = 0.8) +
  scale_fill_manual(values = palette_2) +
  labs(
    x = expression(K[J]),
    y = "Probability",
    title = "Cluster Count Distribution"
  ) +
  theme_minimal() +
  theme(legend.position = "bottom", legend.title = element_blank())

p2 <- ggplot(w1_compare_df, aes(x = w1, y = Density, color = Prior)) +
  geom_line(linewidth = 1) +
  geom_vline(xintercept = 0.5, linetype = "dashed", color = "gray50") +
  scale_color_manual(values = palette_2) +
  labs(
    x = expression(w[1]),
    y = "Density",
    title = "First Size-Biased Weight Distribution"
  ) +
  theme_minimal() +
  theme(legend.position = "bottom", legend.title = element_blank())

gridExtra::grid.arrange(p1, p2, ncol = 2)
```

The dual-anchor prior shifts mass in the $W_{SB}$ distribution away from
extreme values (right panel) while maintaining a similar expected number
of clusters (left panel). This reduces the named tail probability without
substantially changing our beliefs about cluster count.

### 1.8 Sensitivity Analysis

Given the uncertainty in our prior specification, we conduct a sensitivity 
analysis across a range of plausible values for both the K-only and 
dual-anchor approaches:

```{r cct-sensitivity}
# Sensitivity to μ_K
mu_K_grid <- c(3, 5, 7, 9)

sensitivity_results <- lapply(mu_K_grid, function(mu) {
  # K-only fit
  target_k <- DPprior_target_K(J = J_cct, mu_K = mu, confidence = "low")
  fit_k <- DPprior_fit(
    J = J_cct, target_K = target_k, method = "A2-MN", M = 160L,
    check_diagnostics = FALSE
  )
  diag_k <- DPprior_diagnostics(fit_k, allow_approximate = TRUE)
  
  # Verified hard dual-anchor fit
  fit_d <- DPprior_dual_hard(
    fit = fit_k,
    constraint = list(
      metric = "wsb_tail", threshold = 0.5,
      relation = "<=", bound = 0.25
    )
  )
  K_k <- diag_k$diagnostics$K
  K_d <- fit_d$achieved$K
  
  data.frame(
    mu_K = mu,
    # K-only results
    a_k = round(fit_k$parameters$a, 3),
    b_k = round(fit_k$parameters$b, 3),
    E_K_k = round(K_k$mean, 2),
    P_WSB_k = round(prob_wsb_exceeds(
      0.5, fit_k$parameters$a, fit_k$parameters$b
    ), 3),
    # Dual-anchor results
    a_d = round(fit_d$parameters$a, 3),
    b_d = round(fit_d$parameters$b, 3),
    E_K_d = round(K_d$mean, 2),
    P_WSB_d = round(fit_d$achieved$weight$value, 3)
  )
})

sensitivity_df <- do.call(rbind, sensitivity_results)

knitr::kable(
  sensitivity_df,
  col.names = c("Target μ_K", 
                "a (K)", "b (K)", "E[K]", "P(W_SB>0.5)",
                "a (DA)", "b (DA)", "E[K]", "P(W_SB>0.5)"),
  row.names = FALSE,
  caption = "Sensitivity analysis: K-only (K) vs Dual-anchor (DA) priors (J = 38, low confidence)"
)
```

### 1.9 Final Prior Selection and Reporting

Based on our analysis, we recommend the **dual-anchor prior** for the
CCT study, as it provides appropriate control over the predeclared
$W_{SB}$ tail probability in this low between-site information setting.

```{r cct-final}
cat("RECOMMENDED PRIOR FOR CCT STUDY\n")
cat("================================\n")
cat("α ~ Gamma(", round(fit_cct_dual$parameters$a, 4), ", ",
    round(fit_cct_dual$parameters$b, 4), ")\n\n", sep = "")
cat("Key properties:\n")
cat("  E[K] =", round(K_cct_dual$mean, 2), "clusters\n")
cat("  P(W_SB > 0.5) =",
    round(wsb_tail_cct_dual, 2), "\n")
cat("  Hard constraint verified:", fit_cct_dual$constraint$satisfied, "\n")
```

**Reporting language for publication:**

> We elicited beliefs about the number of latent clusters among the 
> $J = 38$ sites in the CCT multisite trial. Based on substantive knowledge 
> of similar education interventions and the uniformity of implementation, 
> we specified $\mathbb{E}[K_J] = 5$ (approximately five distinct response 
> patterns) with low confidence. Initial diagnostics using the K-only 
> calibration yielded $P(W_{SB} > 0.5) = `r round(wsb_tail_cct["threshold_0.5"], 2)`$,
> which is problematic in settings with limited between-site information. 
> We therefore applied verified hard dual-anchor refinement with
> $P(W_{SB} > 0.5) \leq 0.25$, yielding a
> Gamma(`r round(fit_cct_dual$parameters$a, 3)`,
> `r round(fit_cct_dual$parameters$b, 3)`) hyperprior. The refined prior maintains
> $\mathbb{E}[K_J] \approx `r round(K_cct_dual$mean, 1)`$ while
> independently verifying the declared weight inequality. Sensitivity
> analyses across $\mu_K \in \{3, 5, 7, 10\}$ confirmed that substantive 
> conclusions were robust to prior specification.


## 2. Case Study: Brief Alcohol Interventions Meta-Analysis

### 2.1 Research Context

Our second case study is based on the meta-analysis of brief alcohol 
interventions (BAI) for adolescents and young adults conducted by 
[Pustejovsky & Tipton (2022)](https://doi.org/10.1007/s11121-021-01246-3), 
who re-analyzed data from [Tanner-Smith & Lipsey (2015)](https://doi.org/10.1016/j.jsat.2014.09.001).

**Study characteristics:**

- **Number of studies**: 117 randomized trials
- **Total effect sizes**: 1,198 estimates
- **Effect sizes per study**: Median = 6, range = 1 to 108
- **Outcome**: Alcohol consumption (standardized mean differences)
- **Key feature**: Complex dependence structure (correlated and hierarchical)

### 2.2 The Meta-Analytic Challenge

In meta-analysis, the Dirichlet Process mixture model can be used to 
flexibly model the distribution of true study effects. This is particularly 
valuable when:

- The true effect distribution may be non-Gaussian
- There may be discrete subpopulations of studies with similar effects
- Robust inference is needed without strong distributional assumptions

The BAI meta-analysis presents a setting with substantial between-study 
heterogeneity ($\hat{\tau} = 0.182$ under the correlated effects model), 
suggesting the potential for distinct effect subgroups.

### 2.3 Eliciting the Prior

**Step 1: Substantive reasoning about clusters**

In the BAI context, we consider potential sources of effect heterogeneity:

- **Intervention type**: Brief motivational interviewing vs. feedback-only 
  vs. multi-session programs vs. computerized interventions
- **Population**: College students vs. high school vs. clinical samples 
  vs. community populations
- **Outcome timing**: Immediate vs. short-term (3 months) vs. medium-term 
  (6 months) vs. long-term (12+ months) effects
- **Control condition**: No treatment vs. attention control vs. 
  treatment-as-usual vs. alternative intervention
- **Delivery format**: Individual vs. group vs. self-administered
- **Setting**: Healthcare vs. educational vs. community

Given the many potential mechanisms for heterogeneity and the large number 
of studies, a reasonable expectation might be 15--25 distinct effect 
clusters among 117 studies. We start with $\mu_K = 18$:

```{r bai-step1}
# Define the meta-analysis context
J_bai <- 117
mu_K_bai <- 18

cat("Meta-analysis context:\n")
cat("  Number of studies (J):", J_bai, "\n")
cat("  Expected clusters (μ_K):", mu_K_bai, "\n")
cat("  Ratio (μ_K/J):", round(mu_K_bai / J_bai, 3), "\n")
```

**Step 2: Expressing uncertainty**

Meta-analyses often have more uncertainty about the number of distinct 
effect clusters than multisite trials, as studies are conducted 
independently with varying methodologies. We use low confidence:

```{r bai-step2}
target_bai <- DPprior_target_K(
  J = J_bai, mu_K = mu_K_bai, confidence = "low"
)
fit_bai <- DPprior_fit(
  J = J_bai,
  target_K = target_bai,
  method = "A2-MN",
  M = 160L,
  check_diagnostics = FALSE
)

print(fit_bai)
```

**Step 3: Diagnostic verification**

```{r bai-step3}
diag_bai <- DPprior_diagnostics(fit_bai, allow_approximate = TRUE)
alpha_bai <- diag_bai$diagnostics$alpha
K_bai <- diag_bai$diagnostics$K
wsb_bai <- diag_bai$diagnostics$weights
alpha_sd_bai <- sqrt(fit_bai$parameters$a) / fit_bai$parameters$b
alpha_q_bai <- qgamma(
  c(q5 = 0.05, q95 = 0.95),
  shape = fit_bai$parameters$a, rate = fit_bai$parameters$b
)
K_mode_bai <- which.max(K_bai$pmf)
wsb_tail_bai <- prob_wsb_exceeds(
  0.5, fit_bai$parameters$a, fit_bai$parameters$b
)

cat("Diagnostic Summary:\n")
cat("----------------\n")
cat("Alpha distribution:\n")
cat("  E[α] =", round(alpha_bai$mean, 3), "\n")
cat("  SD(α) =", round(alpha_sd_bai, 3), "\n")
cat("  90% central interval: [", round(alpha_q_bai["q5"], 3), ", ",
    round(alpha_q_bai["q95"], 3), "]\n", sep = "")
cat("\nCluster count:\n")
cat("  E[K] =", round(K_bai$mean, 3), "\n")
cat("  Var(K) =", round(K_bai$variance, 3), "\n")
cat("  Mode(K) =", K_mode_bai, "\n")
cat("\nWeight behavior:\n")
cat("  E[W_SB] =", round(wsb_bai$mean, 3), "\n")
cat("  P(W_SB > 0.5) =",
    round(wsb_tail_bai, 3), "\n")
print(diag_bai[c("status", "usable", "verified", "message")])
```

### 2.4 Comparing Alternative Specifications

For meta-analysis, it is valuable to compare different prior specifications 
to assess robustness. We compare three approaches: expecting fewer clusters, 
the baseline, and expecting more clusters.

```{r bai-comparison}
# Define three candidate priors
fit_bai_candidate <- function(mu_K) {
  target_K <- DPprior_target_K(
    J = J_bai, mu_K = mu_K, confidence = "low"
  )
  DPprior_fit(
    J = J_bai, target_K = target_K, method = "A2-MN", M = 160L,
    check_diagnostics = FALSE
  )
}
bai_candidates <- list(
  "Conservative (μ_K = 10)" = fit_bai_candidate(10),
  "Baseline (μ_K = 18)" = fit_bai_candidate(18),
  "Liberal (μ_K = 30)" = fit_bai_candidate(30)
)

# Create comparison table
bai_comp_results <- lapply(names(bai_candidates), function(nm) {
  fit <- bai_candidates[[nm]]
  diag <- DPprior_diagnostics(fit, allow_approximate = TRUE)
  K_diag <- diag$diagnostics$K
  wsb_diag <- diag$diagnostics$weights
  rho_diag <- diag$diagnostics$coclustering
  data.frame(
    Prior = nm,
    a = round(fit$parameters$a, 3),
    b = round(fit$parameters$b, 3),
    E_K = round(K_diag$mean, 2),
    E_w1 = round(wsb_diag$mean, 3),
    E_rho = round(rho_diag$mean, 3),
    Status = diag$status,
    Verified = diag$verified
  )
})

bai_comp_df <- do.call(rbind, bai_comp_results)

knitr::kable(
  bai_comp_df,
  col.names = c(
    "Prior", "a", "b", "E[K]", "E[W_SB]", "E[ρ]",
    "Status", "Verified"
  ),
  caption = "Comparison of candidate priors for the BAI meta-analysis (J = 117)"
)
```

### 2.5 Visualizing the Comparison

```{r bai-viz-comparison, fig.height=6, fig.cap="Comparison of three candidate priors for the BAI meta-analysis."}
# Create data for K PMF comparison
logS_bai <- compute_log_stirling(J_bai)

k_bai_df <- do.call(rbind, lapply(names(bai_candidates), function(nm) {
  fit <- bai_candidates[[nm]]
  pmf <- pmf_K_marginal(
    J_bai, fit$parameters$a, fit$parameters$b, logS = logS_bai
  )
  data.frame(
    K = seq_along(pmf),
    probability = pmf,
    Prior = nm
  )
}))
k_bai_df$Prior <- factor(k_bai_df$Prior, levels = names(bai_candidates))

# Plot PMF comparison
ggplot(k_bai_df[k_bai_df$K <= 50, ], 
       aes(x = K, y = probability, color = Prior)) +
  geom_line(linewidth = 0.8) +
  geom_point(size = 1.5) +
  scale_color_manual(values = palette_3) +
  labs(
    x = expression(K[J]),
    y = "Probability",
    title = "Prior PMF of Number of Clusters (BAI Meta-Analysis)",
    subtitle = "J = 117 studies"
  ) +
  theme_minimal() +
  theme(legend.position = "bottom", legend.title = element_blank())
```

## 3. Case Study: Alabama Pre-K Value-Added Analysis

### 3.1 Research Context

Our third case study illustrates prior elicitation for a large-scale 
value-added analysis in Alabama's pre-kindergarten program.

**Study characteristics:**

- **Sites**: 847 pre-K sites across the state
- **Providers**: 74.4% public schools, 10.4% private centers, 6.2% Head Start, 
  and others
- **Class sizes**: 13--18 students per classroom
- **Outcomes**: Six developmental domains (Literacy, Mathematics, Cognitive, 
  Language, Social-Emotional, Physical)
- **Assessment**: Teaching Strategies GOLD®

**A note on package limitations**: The current version of DPprior supports 
$J \leq 500$ due to pre-computed unsigned Stirling numbers of the first kind, 
which are required for exact moment calculations. Support for larger $J$ 
is planned for future releases. For the 847 sites in the Alabama study, 
we demonstrate the workflow using $J = 500$ as a practical upper bound. 
This limitation has minimal impact on substantive conclusions for several 
reasons: (1) the relationship between $\alpha$ and $\mathbb{E}[K_J]$ is 
approximately logarithmic in $J$, so results are relatively stable across 
nearby sample sizes; (2) using a conservative (smaller) $J$ tends to 
produce slightly more concentrated priors, which is a defensible choice 
in uncertain settings; and (3) the $\mu_K / J$ ratio---rather than $J$ 
alone---is the primary driver of the elicited prior.

### 3.2 The Value-Added Challenge

Value-added models (VAMs) estimate site-specific contributions to 
student learning ($\theta_j$). Key challenges in this context include:

1. **Distribution estimation**: Is the distribution of site effects 
   normal, or are there distinct subpopulations?
2. **Tail identification**: Which sites fall in the lowest and 
   highest 10% of the distribution?
3. **Ranking**: Can we produce meaningful "report cards" for sites?

The semiparametric Dirichlet Process approach allows flexible modeling 
of the site effect distribution without imposing normality.

### 3.3 Eliciting the Prior

**Step 1: Substantive considerations**

With 847 sites, we expect considerable heterogeneity. Potential sources 
of clustering include:

- **Provider type**: Public schools vs. private centers vs. Head Start 
  may exhibit systematically different effects
- **Geographic region**: Urban vs. rural patterns
- **Resource levels**: Varying levels of site resources and teacher 
  qualifications
- **Implementation fidelity**: Variation in how the curriculum is delivered

A reasonable expectation might be 40--60 distinct "types" of sites, 
representing approximately 5--7% of the total. We specify $\mu_K = 50$:

```{r alabama-step1}
# Define the context
# Note: Using J = 500 (package maximum) for the 847-site study
J_alabama <- 500
mu_K_alabama <- 50

cat("Alabama Pre-K context:\n")
cat("  Actual number of sites: 847\n")
cat("  J used for elicitation:", J_alabama, "(package maximum)\n")
cat("  Expected clusters (μ_K):", mu_K_alabama, "\n")
cat("  Ratio (μ_K/J):", round(mu_K_alabama / J_alabama, 3), "\n")
```

**Step 2: Prior elicitation**

```{r alabama-step2}
# Construct the target and fit it using A2-MN
target_alabama <- DPprior_target_K(
  J = J_alabama, mu_K = mu_K_alabama, confidence = "low"
)
fit_alabama <- DPprior_fit(
  J = J_alabama,
  target_K = target_alabama,
  method = "A2-MN",
  M = 160L,
  check_diagnostics = FALSE
)

print(fit_alabama)
```

**Step 3: Diagnostics**

```{r alabama-diagnostics}
diag_alabama <- DPprior_diagnostics(
  fit_alabama, allow_approximate = TRUE
)
alpha_alabama <- diag_alabama$diagnostics$alpha
K_alabama <- diag_alabama$diagnostics$K
wsb_alabama <- diag_alabama$diagnostics$weights
alpha_sd_alabama <- sqrt(fit_alabama$parameters$a) /
  fit_alabama$parameters$b
alpha_q_alabama <- qgamma(
  c(q5 = 0.05, q95 = 0.95),
  shape = fit_alabama$parameters$a, rate = fit_alabama$parameters$b
)
wsb_tail_alabama <- prob_wsb_exceeds(
  0.5, fit_alabama$parameters$a, fit_alabama$parameters$b
)

cat("Diagnostic Summary:\n")
cat("----------------\n")
cat("Alpha distribution:\n")
cat("  E[α] =", round(alpha_alabama$mean, 3), "\n")
cat("  SD(α) =", round(alpha_sd_alabama, 3), "\n")
cat("  90% central interval: [", round(alpha_q_alabama["q5"], 3), ", ",
    round(alpha_q_alabama["q95"], 3), "]\n", sep = "")
cat("\nCluster count:\n")
cat("  E[K] =", round(K_alabama$mean, 2), "\n")
cat("  SD(K) =", round(sqrt(K_alabama$variance), 2), "\n")
cat("\nWeight behavior:\n")
cat("  E[W_SB] =", round(wsb_alabama$mean, 4), "\n")
cat("  P(W_SB > 0.5) =",
    round(wsb_tail_alabama, 4), "\n")
print(diag_alabama[c("status", "usable", "verified", "message")])
```

### 3.4 Sensitivity Analysis

We examine how the elicited prior changes across different specifications:

```{r alabama-sensitivity}
# Sensitivity to μ_K
mu_K_grid <- c(30, 40, 50, 60, 75)

sensitivity_alabama <- lapply(mu_K_grid, function(mu) {
  target_K <- DPprior_target_K(
    J = J_alabama, mu_K = mu, confidence = "low"
  )
  fit <- DPprior_fit(
    J = J_alabama, target_K = target_K, method = "A2-MN", M = 160L,
    check_diagnostics = FALSE
  )
  diag <- DPprior_diagnostics(fit, allow_approximate = TRUE)
  K_diag <- diag$diagnostics$K
  data.frame(
    mu_K = mu,
    ratio = round(mu / J_alabama, 3),
    a = round(fit$parameters$a, 3),
    b = round(fit$parameters$b, 3),
    E_K = round(K_diag$mean, 1),
    P_WSB = round(prob_wsb_exceeds(
      0.5, fit$parameters$a, fit$parameters$b
    ), 4)
  )
})

sensitivity_alabama_df <- do.call(rbind, sensitivity_alabama)

knitr::kable(
  sensitivity_alabama_df,
  col.names = c("Target μ_K", "μ_K/J", "a", "b", "E[K]", "P(W_SB > 0.5)"),
  caption = "Sensitivity analysis for the Alabama Pre-K study (J = 500)"
)
```

### 3.5 Visualizing Prior Behavior

```{r alabama-viz, fig.height=5, fig.cap="Prior distributions for the Alabama Pre-K study (J = 500, μ_K = 50)."}
# Plot alpha and K distributions side by side
# Alpha distribution
alpha_grid <- seq(0.01, 25, length.out = 300)
alpha_df <- data.frame(
  alpha = alpha_grid,
  density = dgamma(
    alpha_grid,
    shape = fit_alabama$parameters$a,
    rate = fit_alabama$parameters$b
  )
)

p1 <- ggplot(alpha_df, aes(x = alpha, y = density)) +
  geom_line(linewidth = 1, color = "#377EB8") +
  geom_vline(xintercept = alpha_alabama$mean,
             linetype = "dashed", color = "gray40") +
  annotate("text", x = alpha_alabama$mean + 1.5,
           y = max(alpha_df$density) * 0.9,
           label = sprintf("E[α] = %.1f", alpha_alabama$mean),
           hjust = 0, color = "gray40") +
  labs(
    x = expression(alpha),
    y = "Density",
    title = expression("Prior on " * alpha)
  ) +
  theme_minimal()

# K distribution
logS_al <- compute_log_stirling(J_alabama)
pmf_K <- pmf_K_marginal(
  J_alabama, fit_alabama$parameters$a, fit_alabama$parameters$b,
  logS = logS_al
)
k_df <- data.frame(
  K = seq_along(pmf_K),
  probability = pmf_K
)

p2 <- ggplot(k_df[k_df$K <= 100 & k_df$probability > 1e-4, ], 
             aes(x = K, y = probability)) +
  geom_col(fill = "#377EB8", alpha = 0.7) +
  geom_vline(xintercept = K_alabama$mean,
             linetype = "dashed", color = "gray40") +
  annotate("text", x = K_alabama$mean + 5,
           y = max(k_df$probability) * 0.9,
           label = sprintf("E[K] = %.0f", K_alabama$mean),
           hjust = 0, color = "gray40") +
  labs(
    x = expression(K[J]),
    y = "Probability",
    title = expression("Prior PMF of " * K[J])
  ) +
  theme_minimal()

gridExtra::grid.arrange(p1, p2, ncol = 2)
```


## 4. Reporting Prior Elicitation in Publications

### 4.1 Key Elements to Report

When reporting prior elicitation in a publication, include:

1. **Study context**: Number of units ($J$) and their nature
2. **Substantive rationale**: Why you expect a certain number of clusters
3. **Uncertainty characterization**: Confidence level or variance specification
4. **Algorithm used**: A1, A2-MN, or A2-KL
5. **Resulting hyperparameters**: The $\text{Gamma}(a, b)$ specification
6. **Diagnostic summary**: Named quantities such as $P(W_{SB} > 0.5)$,
   separately named certified $W_{max}$ upper-bound evidence only when its
   typed method contract authorizes it, and the
   `status`/`usable`/`verified`/`message` quartet
7. **Sensitivity analysis**: Results for alternative specifications



## Summary

This vignette demonstrated how to apply the DPprior package across three 
diverse research contexts:

| Case Study | J | μ_K | Key Insight |
|------------|---|-----|-------------|
| CCT Multisite Trial | 38 | 5 | Dual-anchor refinement essential for low-information settings |
| BAI Meta-Analysis | 117 | 18 | Multiple heterogeneity sources suggest more clusters |
| Alabama Pre-K VAM | 500 | 50 | Large-scale applications work within package limits |

**Key Takeaways:**

1. **Context matters**: The appropriate $\mu_K$ depends on substantive 
   domain knowledge about heterogeneity sources
   
2. **Diagnostics are essential**: Always verify weight behavior, not just 
   cluster counts
   
3. **Dual-anchor refinement**: When a named weight tail is unacceptable, use
   `DPprior_dual_hard()` for a genuine inequality or `DPprior_dual_soft()`
   for an explicit fixed-scale trade-off

4. **Sensitivity analysis is mandatory**: Report results across plausible 
   alternative specifications
   
5. **Transparent reporting**: Document the complete elicitation process 
   in publications


## References

Barrera-Osorio, F., Linden, L. L., & Saavedra, J. E. (2019). Medium- and 
long-term educational consequences of alternative conditional cash transfer 
designs: Experimental evidence from Colombia. *American Economic Journal: 
Applied Economics*, 11(3), 54–91. https://doi.org/10.1257/app.20170008

Lee, J., Che, J., Rabe-Hesketh, S., Feller, A., & Miratrix, L. (2025). 
Improving the estimation of site-specific effects and their distribution 
in multisite trials. *Journal of Educational and Behavioral Statistics*, 
50(5), 731–764. https://doi.org/10.3102/10769986241254286

Pustejovsky, J. E., & Tipton, E. (2022). Meta-analysis with robust variance 
estimation: Expanding the range of working models. *Prevention Science*, 
23(3), 425–438. https://doi.org/10.1007/s11121-021-01246-3

Tanner-Smith, E. E., & Lipsey, M. W. (2015). Brief alcohol interventions 
for adolescents and young adults: A systematic review and meta-analysis. 
*Journal of Substance Abuse Treatment*, 51, 1–18. 
https://doi.org/10.1016/j.jsat.2014.09.001

---

*For questions or feedback, please visit the 
[GitHub repository](https://github.com/joonho112/DPprior).*
