---
title: "Diagnosing Your Prior: Avoiding Unintended Consequences"
author: "JoonHo Lee (jlee296@ua.edu)"
date: "`r Sys.Date()`"
description: >
  Learn to verify that your elicited Gamma hyperprior behaves as intended 
  across all relevant dimensions, not just the number of clusters.
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Diagnosing Your Prior: Avoiding Unintended Consequences}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
library(DPprior)
library(ggplot2)
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5,
  fig.align = "center",
  out.width = "85%"
)
set.seed(42)

# Define a consistent color palette
palette_3 <- c("#E41A1C", "#377EB8", "#4DAF4A")
palette_4 <- c("#E41A1C", "#377EB8", "#4DAF4A", "#984EA3")
```

## Overview

When you elicit a Gamma hyperprior for $\alpha$ based on your expectations 
about the number of clusters $K_J$, you are implicitly specifying priors 
on *many other quantities* that you may not have consciously considered. 
These "unintended priors" can lead your posterior inference in unexpected 
directions, particularly in low-information settings where the data cannot 
overwhelm an inadvertently strong prior.

This vignette introduces the diagnostic tools in the DPprior package that 
help you verify your prior behaves as intended across all relevant 
dimensions. By the end, you will understand:

1. Why diagnostics matter: the "unintended prior" problem
2. The complete diagnostic suite available in DPprior
3. How to interpret $\alpha$ and weight distribution diagnostics
4. Explicit estimand-specific warning policies and how to respond to them
5. How to use diagnostics to iteratively refine your prior


## 1. Why Diagnostics Matter

### 1.1 The Blind Spot in Prior Elicitation

The standard workflow for eliciting a DP prior focuses on cluster counts:

1. You think about $\mu_K$: "How many groups do I expect?"
2. You think about $\text{Var}(K)$: "How uncertain am I about that?"
3. You build `DPprior_target_K()`, pass it to `DPprior_fit()`, and obtain
   $\alpha \sim \text{Gamma}(a, b)$

This is intuitive and principled. However, your elicited prior on $\alpha$ 
determines not just $K_J$, but also:
 
- The **distribution of $\alpha$ itself**: mean, variance, and shape
- The **cluster weight distribution**: how mass is allocated across clusters
- The **co-clustering probability**: how likely two random observations share a cluster
- A named **weight-tail probability**, such as $P(W_{SB}>0.5)$, kept
  distinct from largest-weight evidence for $W_{max}$

These characteristics are implicit consequences of your $(a, b)$ choice. 
You may not have intended any particular behavior for these quantities, 
yet your prior assigns specific probabilities to them.

### 1.2 The Unintended Prior Problem

[Vicentini & Jermyn (2025)](https://doi.org/10.48550/arXiv.2502.00864) 
demonstrated that matching a target $K_J$ distribution does not guarantee 
intuitive weight behavior. Consider a concrete example:

```{r unintended-demo}
J <- 50
mu_K <- 5

# Canonical K-only calibration
target_K <- DPprior_target_K(J = J, mu_K = mu_K, var_K = 8)
fit_K <- DPprior_fit(J = J, target_K = target_K, method = "A2-MN")

cat("K-only prior: Gamma(", round(fit_K$parameters$a, 3), ", ",
    round(fit_K$parameters$b, 3), ")\n", sep = "")
cat("\nCluster count behavior:\n")
cat("  E[K] =", round(fit_K$achieved$K$mean, 2), "\n")
cat("  This matches our target of", mu_K, "clusters.\n")

cat("\nBut what about weight behavior?\n")
cat("  E[w₁] =",
    round(mean_w1(fit_K$parameters$a, fit_K$parameters$b), 3), "\n")
cat("  P(W_SB > 0.3) =",
    round(prob_wsb_exceeds(0.3, fit_K$parameters$a, fit_K$parameters$b), 3), "\n")
cat("  P(W_SB > 0.5) =",
    round(prob_wsb_exceeds(0.5, fit_K$parameters$a, fit_K$parameters$b), 3), "\n")
cat("  P(W_SB > 0.9) =",
    round(prob_wsb_exceeds(0.9, fit_K$parameters$a, fit_K$parameters$b), 3), "\n")
fit_K[c("status", "usable", "verified", "message")]
```

The prior expects 5 clusters, but there is a substantial probability 
(nearly 50%) that a randomly selected observation belongs to a cluster 
containing more than half of all observations. Is this what you intended 
when you said "I expect 5 clusters"?

Most researchers, when they think "5 clusters," imagine something like 
five roughly comparable groups---not a situation where one cluster might 
dominate the entire mixture.

### 1.3 Why This Matters in Practice

The unintended prior problem is especially consequential in 
**low-information settings** ([Lee et al., 2025](https://doi.org/10.3102/10769986241254286)):

- **Multisite trials with few sites**: When $J$ is moderate (20--100), 
  each site provides limited data
- **Sparse clustering applications**: When observations per cluster are small
- **Exploratory analyses**: When you don't have strong prior data to inform expectations

In these settings, the posterior can remain close to the prior. If your 
prior places more mass than intended on large size-biased clusters, your
posterior may inherit this behavior even when the data suggest otherwise.


## 2. The Complete Diagnostic Suite

The DPprior package provides a comprehensive set of diagnostics that you 
can access in several ways.

### 2.1 Automatic Diagnostics During Fitting

The simplest approach is to enable diagnostics during the fitting process:

```{r diagnostics-during-fit}
target_K <- DPprior_target_K(J = 50, mu_K = 5, var_K = 8)
fit <- DPprior_fit(
  J = 50, target_K = target_K, check_diagnostics = TRUE
)
```

When `check_diagnostics = TRUE`, the function computes all diagnostic 
quantities and stores them in the returned object.

### 2.2 Post-Hoc Diagnostic Computation

You can also compute diagnostics after fitting:
 
```{r posthoc-diagnostics}
# Fit without diagnostics first
target_K <- DPprior_target_K(J = 50, mu_K = 5, var_K = 8)
fit <- DPprior_fit(
  J = 50, target_K = target_K, check_diagnostics = FALSE
)

# Then compute and verify diagnostics. W_SB evidence is named explicitly;
# a W_max policy remains explicitly indeterminate when its backend is unavailable.
diag <- DPprior_diagnostics(fit)
print(diag)
diag[c("status", "usable", "verified", "message")]
```

### 2.3 Overview of Diagnostic Components

The diagnostic object contains several components:

| Component | Description |
|-----------|-------------|
| `diagnostics$alpha` | Verified $\alpha$ mean and CV |
| `diagnostics$K` | Verified $K_J$ mean, variance, PMF, and quadrature order |
| `diagnostics$weights` | Verified first-size-biased-weight mean $E[W_{SB}]$ |
| `diagnostics$coclustering` | Verified co-clustering mean and variance |
| `diagnostics$policy_results` | Exact $W_{SB}$ policy evidence, or an explicit indeterminate $W_{max}$ record when its backend is unavailable |
| `diagnostics$warnings` | One warning for each triggered named policy |

Let us examine each component in detail.


## 3. Alpha Distribution Diagnostics

The first set of diagnostics concerns the concentration parameter $\alpha$ 
itself. Since $\alpha \sim \text{Gamma}(a, b)$, we can compute all moments 
analytically.

### 3.1 Accessing Alpha Diagnostics

```{r alpha-diagnostics}
target_K <- DPprior_target_K(J = 50, mu_K = 5, var_K = 8)
fit <- DPprior_fit(
  J = 50, target_K = target_K, check_diagnostics = FALSE
)
diag <- DPprior_diagnostics(fit)
alpha_diag <- diag$diagnostics$alpha
alpha_sd <- sqrt(fit$parameters$a) / fit$parameters$b
alpha_quantiles <- qgamma(
  c(q5 = 0.05, q25 = 0.25, q50 = 0.50, q75 = 0.75, q95 = 0.95),
  shape = fit$parameters$a, rate = fit$parameters$b
)

cat("Alpha distribution summary:\n")
cat("  Mean:     E[α] =", round(alpha_diag$mean, 4), "\n")
cat("  SD:       SD(α) =", round(alpha_sd, 4), "\n")
cat("  CV:       CV(α) =", round(alpha_diag$CV, 4), "\n")
cat("\nQuantiles:\n")
cat("   5th percentile:",
    round(alpha_quantiles["q5"], 4), "\n")
cat("  25th percentile:",
    round(alpha_quantiles["q25"], 4), "\n")
cat("  50th percentile:",
    round(alpha_quantiles["q50"], 4), "\n")
cat("  75th percentile:",
    round(alpha_quantiles["q75"], 4), "\n")
cat("  95th percentile:",
    round(alpha_quantiles["q95"], 4), "\n")
```

### 3.2 Interpreting the Coefficient of Variation

The coefficient of variation (CV) of $\alpha$ is particularly informative:

| CV Range | Interpretation |
|----------|----------------|
| CV < 0.3 | **Highly informative**: You are quite certain about $\alpha$ |
| 0.3 ≤ CV < 0.5 | **Moderately informative**: Reasonable uncertainty |
| 0.5 ≤ CV < 1.0 | **Weakly informative**: Substantial uncertainty about $\alpha$ |
| CV ≥ 1.0 | **Highly diffuse**: Very uncertain about $\alpha$ |

For a Gamma distribution, $\text{CV}(\alpha) = 1/\sqrt{a}$, so the CV 
depends only on the shape parameter $a$.

```{r cv-interpretation, fig.height=5, fig.cap="Different levels of informativeness in the α prior, controlled by the shape parameter a."}
# Compare different CV levels
a_values <- c(0.5, 1, 2, 4, 10)
cv_values <- 1 / sqrt(a_values)

alpha_grid <- seq(0.01, 8, length.out = 300)
cv_df <- do.call(rbind, lapply(seq_along(a_values), function(i) {
  a <- a_values[i]
  # Use b = a so E[alpha] = 1 for all
  b <- a
  data.frame(
    alpha = alpha_grid,
    density = dgamma(alpha_grid, shape = a, rate = b),
    CV = sprintf("CV = %.2f (a = %.1f)", cv_values[i], a)
  )
}))
cv_df$CV <- factor(cv_df$CV, levels = unique(cv_df$CV))

ggplot(cv_df, aes(x = alpha, y = density, color = CV)) +
  geom_line(linewidth = 1) +
  scale_color_viridis_d(option = "plasma", end = 0.85) +
  labs(
    x = expression(alpha),
    y = "Density",
    title = "Prior Informativeness Controlled by Shape Parameter",
    subtitle = "All priors have E[α] = 1; higher a means lower CV (more informative)"
  ) +
  theme_minimal() +
  theme(legend.position = "right")
```


## 4. Weight Distribution Diagnostics

The weight diagnostics are crucial for detecting unintended prior behavior. 
The DPprior package provides comprehensive diagnostics for the first
size-biased stick $W_{SB}$ (historically named `w1` in helper functions).

### 4.1 Understanding $W_{SB}$

Recall from the [Dual-Anchor vignette](dual-anchor.html) that under 
Sethuraman's stick-breaking representation:

$$
w_1 = v_1, \quad v_1 \mid \alpha \sim \text{Beta}(1, \alpha)
$$

The quantity $w_1=W_{SB}$ has a natural interpretation: it is the **proportion
of the cluster containing a randomly selected observation**. If 
$P(w_1 > 0.5)$ is high, a random observation is likely to belong to a 
cluster containing more than half of all observations.

### 4.2 Accessing Weight Diagnostics

```{r weight-diagnostics}
wsb <- diag$diagnostics$weights
wsb_median <- quantile_w1(
  0.5, fit$parameters$a, fit$parameters$b
)
wsb_tail <- vapply(
  c(0.5, 0.9), prob_wsb_exceeds, numeric(1),
  a = fit$parameters$a, b = fit$parameters$b
)
names(wsb_tail) <- c("threshold_0.5", "threshold_0.9")
cat("Weight distribution summary:\n")
cat("  E[W_SB] =", round(wsb$mean, 4), "\n")
cat("  Median(W_SB) =", round(wsb_median, 4), "\n")

cat("\nNamed W_SB tail probabilities from the closed-form helper:\n")
print(wsb_tail)
cat("\nW_SB diagnostic status fields:\n")
print(wsb[c("status", "usable", "verified", "message")])
```

Canonical diagnostics do not substitute a $W_{SB}$ value for $W_{max}$. A
$W_{max}$ warning policy is retained as `outcome = "indeterminate"` with
`basis = "backend_unavailable"` unless an independently authorized backend
is available. Certified upper-bound constraints belong to the separate
`wmax_tail_upper` hard-dual route.

### 4.3 Closed-Form Calculations

All $w_1$ diagnostics are computed using closed-form expressions (derived
in [Lee, 2026](https://arxiv.org/abs/2602.06301), following [Vicentini & Jermyn, 2025](https://doi.org/10.48550/arXiv.2502.00864)):

**CDF:**
$$
F_{w_1}(x \mid a, b) = 1 - \left( \frac{b}{b - \log(1-x)} \right)^a
$$

**Quantile function:**
$$
Q_{w_1}(u \mid a, b) = 1 - \exp\left( b \left[ 1 - (1-u)^{-1/a} \right] \right)
$$

**First-size-biased-weight tail probability:**
$$
P(w_1 > t \mid a, b) = \left( \frac{b}{b - \log(1-t)} \right)^a
$$

You can access these functions directly:

```{r closed-form-demo}
a <- fit$parameters$a
b <- fit$parameters$b

# Direct computation of tail probabilities
cat("Direct computation of P(W_SB > t):\n")
for (t in c(0.3, 0.5, 0.7, 0.9)) {
  p <- prob_wsb_exceeds(t, a, b)
  cat(sprintf("  P(W_SB > %.1f) = %.4f\n", t, p))
}

# Quantiles
cat("\nQuantiles of W_SB distribution:\n")
for (q in c(0.25, 0.5, 0.75, 0.95)) {
  x <- quantile_w1(q, a, b)
  cat(sprintf("  %d%% quantile: %.4f\n", round(100*q), x))
}
```


## 5. Explicit Estimand-Specific Warning Policies

DPprior does not assign an ambiguous LOW/MODERATE/HIGH category. A warning
policy must name the estimand (`"W_SB"` or `"W_max"`), mass threshold,
comparison direction, and action threshold.

### 5.1 Separate the Mass Threshold from the Action Threshold

For example, the policy below asks whether
$P(W_{SB}>0.5)$ exceeds 0.40. The two 0.5 and 0.40 values answer different
questions and are recorded separately.

```r
warning_policy <- list(
  estimand = "W_SB",
  direction = "above",
  weight_threshold = 0.5,
  action_threshold = 0.40
)
```

### 5.2 Enabling an Explicit Warning Policy

You can enable automatic warnings during fitting:

```{r dominance-warning-demo}
# A prior with a predeclared W_SB policy
target_risky <- DPprior_target_K(J = 50, mu_K = 2, var_K = 2)
fit_risky <- DPprior_fit(
  J = 50, target_K = target_risky,
  warning_policy = list(
    estimand = "W_SB", direction = "above",
    weight_threshold = 0.5, action_threshold = 0.40
  )
)
```

The warning reports the named estimand and exact policy comparison. It does
not claim that $W_{SB}$ is the largest weight.

### 5.3 Reviewing the Policy Record

You can also review the named policy record in the diagnostic object:

```{r dominance-check}
diag_risky <- DPprior_diagnostics(
  fit_risky,
  warning_policy = list(
    estimand = "W_SB", direction = "above",
    weight_threshold = 0.5, action_threshold = 0.40
  )
)

cat("Named policy record:\n")
policy_record <- diag_risky$diagnostics$policy_results[[1]]
print(policy_record)
cat("  P(W_SB > 0.5) =",
    round(policy_record$value, 4), "\n")
```

### 5.4 Interpreting a Triggered Policy

A triggered policy means the named comparison was met. For the example above,
the exact $W_{SB}$ tail probability exceeded the predeclared 0.40 action
threshold.
This may or may not be appropriate depending on your application:

**When a large $W_{SB}$ tail probability might be appropriate:**

- You genuinely expect most data to fall into one dominant group
- You are modeling rare events where most observations are "baseline"
- You have strong prior evidence of an unbalanced partition

**When a large $W_{SB}$ tail probability is likely unintended:**

- You expect roughly equal-sized clusters
- You think "$K$ clusters" should mean roughly balanced groups
- You are doing exploratory clustering without strong prior expectations

### 5.5 Responding to a Triggered Policy

If a predeclared weight policy triggers and that behavior is unintended, you
have several options:

1. **Increase $\mu_K$**: Expecting more clusters typically reduces the
   $W_{SB}$ tail probability
2. **Decrease $\text{Var}(K)$**: More certainty about cluster count can help
3. **Use dual-anchor elicitation**: Directly constrain weight behavior

We will demonstrate option 3 in Section 7.


## 6. Visualizing Diagnostics

The DPprior package provides several visualization functions for 
understanding your prior.

### 6.1 The Complete Dashboard

The `plot()` method for a `DPprior_fit` object creates a comprehensive 
four-panel dashboard:

```{r dashboard, fig.height=8, fig.cap="Complete diagnostic dashboard showing all key prior characteristics."}
target_K <- DPprior_target_K(J = 50, mu_K = 5, var_K = 8)
fit <- DPprior_fit(
  J = 50, target_K = target_K, check_diagnostics = FALSE
)
plot(fit)
```

### 6.2 Individual Diagnostic Plots

For more focused analysis, you can create individual plots:

```{r fig-alpha-code, fig.cap="Prior density for the concentration parameter α.", eval=FALSE}
# Alpha prior density
plot_alpha_prior(fit)
```

```{r fig-alpha-plot, echo=FALSE, fig.height=4, fig.cap="Prior density for the concentration parameter alpha.", fig.alt="A smooth density curve showing the Gamma prior distribution on the concentration parameter alpha."}
invisible(plot_alpha_prior(fit))
```

```{r fig-K-code, fig.cap="Marginal PMF of the number of clusters K.", eval=FALSE}
# K marginal PMF
plot_K_prior(fit)
```

```{r fig-K-plot, echo=FALSE, fig.height=4, fig.cap="Marginal PMF of the number of clusters K.", fig.alt="A bar plot showing the prior probability mass over possible numbers of clusters."}
invisible(plot_K_prior(fit))
```

```{r fig-w1-code, fig.cap="Distribution of the first size-biased weight W_SB.", eval=FALSE}
# First-size-biased-weight distribution with reviewed mass thresholds
plot_w1_prior(fit)
```

```{r fig-w1-plot, echo=FALSE, fig.height=4, fig.cap="Distribution of the first size-biased weight W_SB.", fig.alt="A density plot showing the prior distribution of the first size-biased weight and reviewed mass thresholds."}
invisible(plot_w1_prior(fit))
```

### 6.3 Comparing Multiple Priors

When considering different prior specifications, you can visualize them 
side by side:

```{r compare-priors, fig.height=5, fig.cap="Comparison of K distributions under different variance assumptions."}
comparison_targets <- list(
  "Low uncertainty (var_K = 6)" =
    DPprior_target_K(J = 50, mu_K = 5, var_K = 6),
  "Medium uncertainty (var_K = 10)" =
    DPprior_target_K(J = 50, mu_K = 5, var_K = 10),
  "High uncertainty (var_K = 20)" =
    DPprior_target_K(J = 50, mu_K = 5, var_K = 20)
)
fits <- lapply(comparison_targets, function(target_K) {
  DPprior_fit(J = 50, target_K = target_K, check_diagnostics = FALSE)
})

# Compute log Stirling numbers for efficiency
logS <- compute_log_stirling(50)

# Build comparison data
k_df <- do.call(rbind, lapply(names(fits), function(nm) {
  fit <- fits[[nm]]
  pmf <- pmf_K_marginal(
    50, fit$parameters$a, fit$parameters$b, logS = logS
  )
  data.frame(
    K = seq_along(pmf) - 1,  # pmf_K_marginal returns k=0,...,J
    probability = pmf,
    Prior = nm
  )
}))
k_df <- k_df[k_df$K >= 1, ]  # Remove k=0
k_df$Prior <- factor(k_df$Prior, levels = names(fits))

# Plot
ggplot(k_df[k_df$K <= 20, ], aes(x = K, y = probability, color = Prior)) +
  geom_point(size = 2) +
  geom_line(linewidth = 0.8) +
  scale_color_manual(values = palette_3) +
  labs(
    x = expression(Number~of~clusters~K[J]),
    y = "Probability",
    title = "Prior PMF of K Under Different Variance Assumptions",
    subtitle = "All priors have μ_K = 5; higher variance spreads the distribution"
  ) +
  theme_minimal() +
  theme(legend.position = "bottom", legend.title = element_blank())
```


## 7. Diagnostic-Driven Refinement

The diagnostic tools enable an iterative workflow for refining your prior 
until it matches your intentions across all dimensions.

### 7.1 The Refinement Workflow

The recommended workflow is:

1. **Start with K-only calibration**: Build `DPprior_target_K()` and pass it
   unchanged to `DPprior_fit()`
2. **Run diagnostics**: Check $\alpha$, $K_J$, and explicitly named weight estimands
3. **Identify mismatches**: Does any aspect of the prior surprise you?
4. **Refine**: Adjust parameters or use dual-anchor if needed
5. **Repeat**: Until all aspects of the prior match your intentions

### 7.2 Complete Example: From Problem to Solution

Let us walk through a complete example of diagnostic-driven refinement.

**Step 1: Initial K-only prior**

```{r refinement-step1}
# Researcher expects ~5 clusters with moderate uncertainty
target_K <- DPprior_target_K(J = 50, mu_K = 5, var_K = 8)
fit1 <- DPprior_fit(
  J = 50, target_K = target_K, method = "A2-MN",
  check_diagnostics = FALSE
)

cat("Step 1: Initial K-only prior\n")
cat("Gamma(a =", round(fit1$parameters$a, 4), ", b =",
    round(fit1$parameters$b, 4), ")\n\n")
```

**Step 2: Run diagnostics**

```{r refinement-step2}
diag1 <- DPprior_diagnostics(fit1)
K1 <- diag1$diagnostics$K
wsb1 <- diag1$diagnostics$weights
wsb1_tail <- prob_wsb_exceeds(
  0.5, fit1$parameters$a, fit1$parameters$b
)

cat("Step 2: Diagnostics reveal...\n")
cat("  K behavior: E[K] =", round(K1$mean, 2),
    ", Var(K) =", round(K1$variance, 2), "✓\n")
cat("  Weight behavior: P(W_SB > 0.5) =",
    round(wsb1_tail, 3), "\n")
print(diag1[c("status", "usable", "verified", "message")])
```

**Step 3: Identify the problem**

```{r refinement-step3-code, fig.cap="The K-only prior has a larger-than-intended W_SB tail probability.", eval=FALSE}
plot_w1_prior(fit1)
```

```{r refinement-step3, echo=FALSE, fig.height=4, fig.cap="The K-only prior has a larger-than-intended W_SB tail probability.", fig.alt="A diagnostic plot showing substantial first-size-biased-weight mass above 0.5."}
cat("Step 3: Problem identification\n")
cat("  The researcher wanted: ~5 clusters with balanced weights\n")
cat("  The prior implies: ~5 clusters but a large W_SB tail probability\n")
cat("  P(W_SB > 0.5) =",
    round(100 * wsb1_tail),
    "% is higher than expected for 'balanced' clusters\n")

invisible(plot_w1_prior(fit1))
```

**Step 4: Apply dual-anchor refinement**

```{r refinement-step4}
cat("Step 4: Dual-anchor refinement\n")
cat("  Adding constraint: P(W_SB > 0.5) ≤ 0.25\n\n")

fit2 <- DPprior_dual_hard(
  fit = fit1,
  constraint = list(
    metric = "wsb_tail", threshold = 0.5,
    relation = "<=", bound = 0.25
  )
)

cat("Refined prior: Gamma(a =", round(fit2$parameters$a, 4), ", b =",
    round(fit2$parameters$b, 4), ")\n")
cat("Constraint verified:", fit2$constraint$satisfied, "\n")
fit2[c("status", "usable", "verified", "message")]
```

**Step 5: Read the verified hard-fit evidence**

```{r refinement-step5}
K2 <- fit2$achieved$K
wsb2_mean <- mean_w1(fit2$parameters$a, fit2$parameters$b)
wsb2_tail <- fit2$achieved$weight$value

cat("Step 5: Post-refinement evidence\n")
cat("  K behavior: E[K] =", round(K2$mean, 2),
    " (shifted from target of 5)\n")
cat("  Weight behavior: P(W_SB > 0.5) =",
    round(wsb2_tail, 3), "✓\n")
print(fit2[c("status", "usable", "verified", "message")])
```

**Step 6: Visualize the improvement**

```{r refinement-step6, fig.height=5, fig.cap="Comparison of weight distributions before and after dual-anchor refinement."}
# Compare w1 distributions
x_grid <- seq(0.001, 0.999, length.out = 500)
w1_df <- data.frame(
  x = rep(x_grid, 2),
  density = c(
    density_w1(x_grid, fit1$parameters$a, fit1$parameters$b),
    density_w1(x_grid, fit2$parameters$a, fit2$parameters$b)
  ),
  Prior = rep(c("K-only (before)", "Dual-anchor (after)"), each = length(x_grid))
)
w1_df$Prior <- factor(w1_df$Prior, levels = c("K-only (before)", "Dual-anchor (after)"))

# Cap extreme densities for visualization
w1_df$density[w1_df$density > 10] <- NA

ggplot(w1_df, aes(x = x, y = density, color = Prior)) +
  geom_line(linewidth = 1, na.rm = TRUE) +
  geom_vline(xintercept = 0.5, linetype = "dashed", color = "gray50") +
  scale_color_manual(values = c("#E41A1C", "#377EB8")) +
  annotate("text", x = 0.52, y = 5, label = "P(W_SB > 0.5)", hjust = 0, color = "gray40") +
  labs(
    x = expression(w[1]),
    y = "Density",
    title = "Weight Distribution: Before and After Dual-Anchor Refinement",
    subtitle = sprintf("P(W_SB > 0.5): Before = %.0f%%, After = %.0f%%",
                       100 * wsb1_tail,
                       100 * wsb2_tail)
  ) +
  coord_cartesian(xlim = c(0, 1), ylim = c(0, 6)) +
  theme_minimal() +
  theme(legend.position = "bottom", legend.title = element_blank())
```

### 7.3 Summary of the Refinement

```{r refinement-summary}
comparison_df <- data.frame(
  Metric = c(
    "Gamma(a, b)", "E[K]", "Var(K)", "E[W_SB]",
    "P(W_SB > 0.5)", "Diagnostic status"
  ),
  Before = c(
    sprintf("(%.3f, %.3f)", fit1$parameters$a, fit1$parameters$b),
    sprintf("%.2f", K1$mean),
    sprintf("%.2f", K1$variance),
    sprintf("%.3f", wsb1$mean),
    sprintf("%.1f%%", 100 * wsb1_tail),
    diag1$status
  ),
  After = c(
    sprintf("(%.3f, %.3f)", fit2$parameters$a, fit2$parameters$b),
    sprintf("%.2f", K2$mean),
    sprintf("%.2f", K2$variance),
    sprintf("%.3f", wsb2_mean),
    sprintf("%.1f%%", 100 * wsb2_tail),
    fit2$status
  )
)

knitr::kable(
  comparison_df,
  col.names = c("Metric", "K-only", "Dual-anchor"),
  caption = "Comparison of prior specifications before and after dual-anchor refinement"
)
```

The trade-off is clear: the refined prior has slightly higher E[K] than 
the original target of 5, but now has appropriately controlled weight 
behavior. This is the Pareto trade-off that the dual-anchor framework 
makes explicit.


## 8. Comparative Diagnostics

When evaluating multiple candidate priors, comparative diagnostics help 
you understand the trade-offs.

### 8.1 Comparing Multiple Priors

```{r comparative-diagnostics}
# Define three candidate priors
candidate_targets <- list(
  "Conservative" = DPprior_target_K(J = 50, mu_K = 5, var_K = 6),
  "Moderate" = DPprior_target_K(J = 50, mu_K = 5, var_K = 10),
  "Diffuse" = DPprior_target_K(J = 50, mu_K = 5, var_K = 20)
)
candidates <- lapply(candidate_targets, function(target_K) {
  DPprior_fit(J = 50, target_K = target_K, check_diagnostics = FALSE)
})

# Compute diagnostics for each
comp_results <- lapply(names(candidates), function(nm) {
  fit <- candidates[[nm]]
  # Retain a complete review object if fixed-order refinement does not verify.
  # Such a row remains unusable and must not be promoted to a scientific result.
  diag <- DPprior_diagnostics(fit, allow_approximate = TRUE)
  K_diag <- diag$diagnostics$K
  wsb_diag <- diag$diagnostics$weights
  data.frame(
    Prior = nm,
    a = round(fit$parameters$a, 3),
    b = round(fit$parameters$b, 3),
    E_K = round(K_diag$mean, 2),
    Var_K = round(K_diag$variance, 2),
    E_w1 = round(wsb_diag$mean, 3),
    P_WSB_gt_50 = sprintf(
      "%.1f%%", 100 * prob_wsb_exceeds(
        0.5, fit$parameters$a, fit$parameters$b
      )
    ),
    Status = diag$status,
    Usable = diag$usable,
    Verified = diag$verified,
    Message = diag$message
  )
})

comp_df <- do.call(rbind, comp_results)

knitr::kable(
  comp_df,
  col.names = c(
    "Prior", "a", "b", "E[K]", "Var(K)", "E[W_SB]",
    "P(W_SB>0.5)", "Status", "Usable", "Verified", "Message"
  ),
  caption = "Comparative diagnostics for three candidate priors"
)
```

Any row with `usable = FALSE` is retained only for method review; its numerical
summaries are not an approved scientific result.

### 8.2 Visualizing the Comparison

```{r comparative-viz, fig.height=6, fig.cap="Comparison of three candidate priors across multiple dimensions."}
# Create comparison data for w1
w1_comp_df <- do.call(rbind, lapply(names(candidates), function(nm) {
  fit <- candidates[[nm]]
  x <- seq(0.001, 0.999, length.out = 300)
  d <- density_w1(x, fit$parameters$a, fit$parameters$b)
  d[d > 8] <- NA
  data.frame(x = x, density = d, Prior = nm)
}))
w1_comp_df$Prior <- factor(w1_comp_df$Prior, levels = names(candidates))

p1 <- ggplot(w1_comp_df, aes(x = x, y = density, color = Prior)) +
  geom_line(linewidth = 1, na.rm = TRUE) +
  geom_vline(xintercept = 0.5, linetype = "dashed", color = "gray60") +
  scale_color_manual(values = palette_3) +
  coord_cartesian(xlim = c(0, 1), ylim = c(0, 5)) +
  labs(x = expression(w[1]), y = "Density", 
       title = expression("Prior on " * w[1])) +
  theme_minimal() +
  theme(legend.position = "bottom", legend.title = element_blank())

# Create comparison data for K
logS <- compute_log_stirling(50)
k_comp_df <- do.call(rbind, lapply(names(candidates), function(nm) {
  fit <- candidates[[nm]]
  pmf <- pmf_K_marginal(
    50, fit$parameters$a, fit$parameters$b, logS = logS
  )
  data.frame(k = seq_along(pmf) - 1, pmf = pmf, Prior = nm)
}))
k_comp_df <- k_comp_df[k_comp_df$k >= 1, ]
k_comp_df$Prior <- factor(k_comp_df$Prior, levels = names(candidates))

p2 <- ggplot(k_comp_df[k_comp_df$k <= 15, ], 
             aes(x = k, y = pmf, color = Prior)) +
  geom_point(size = 2) +
  geom_line(linewidth = 0.8) +
  scale_color_manual(values = palette_3) +
  labs(x = expression(K[J]), y = "Probability", 
       title = expression("Prior PMF of " * K[J])) +
  theme_minimal() +
  theme(legend.position = "bottom", legend.title = element_blank())

gridExtra::grid.arrange(p1, p2, ncol = 2)
```


## Summary

| Diagnostic Aspect | What to Check | Concern If... |
|-------------------|---------------|---------------|
| $\alpha$ CV | `diag$diagnostics$alpha$CV` | CV > 1 (very diffuse) or CV < 0.3 (very tight) without intention |
| E[K] | `diag$diagnostics$K$mean` | Differs substantially from your target $\mu_K$ |
| Var(K) | `diag$diagnostics$K$variance` | Differs substantially from your target or implied variance |
| P($W_{SB}$ > 0.5) | Explicit `warning_policy` record or `prob_wsb_exceeds()` | Exceeds your predeclared action threshold |
| $W_{max}$ policy | `diag$diagnostics$policy_results[[1]]` | An indeterminate/backend-unavailable result is reported as a point claim |
| Decision status | `diag[c("status","usable","verified","message")]` | Approximation or failed verification is hidden |

**Key Takeaways:**

1. **Always run diagnostics** after eliciting a prior, especially for 
   weight-related quantities
2. **K-calibration is necessary but not sufficient**: matching E[K] does 
   not guarantee intuitive weight behavior
3. **Use an explicit estimand-specific warning policy** to catch potential
   problems early
4. **Apply `DPprior_dual_hard()`** when you need a verified inequality, or
   `DPprior_dual_soft()` when you intend a fixed-scale trade-off
5. **Iterate until satisfied**: prior elicitation is an iterative process


## What's Next?

+ **[Dual-Anchor Framework](dual-anchor.html)**: Deep dive into controlling 
  both cluster counts and weight behavior
  
+ **[Case Studies](case-studies.html)**: Real-world examples of 
  diagnostic-driven prior refinement
  
+ **[Theory Overview](theory-overview.html)**: Mathematical foundations 
  of the diagnostic quantities


## References

Lee, J., Che, J., Rabe-Hesketh, S., Feller, A., & Miratrix, L. (2025). 
Improving the estimation of site-specific effects and their distribution 
in multisite trials. *Journal of Educational and Behavioral Statistics*, 
50(5), 731–764. https://doi.org/10.3102/10769986241254286

Vicentini, S., & Jermyn, I. H. (2025). Prior selection for the precision 
parameter of Dirichlet process mixture models. *arXiv:2502.00864*. 
https://doi.org/10.48550/arXiv.2502.00864

Sethuraman, J. (1994). A constructive definition of Dirichlet priors. 
*Statistica Sinica*, 4(2), 639–650.

---

*For questions or feedback, please visit the 
[GitHub repository](https://github.com/joonho112/DPprior).*
