---
title: "Beyond Cluster Counts: Dual-Anchor Elicitation for Weight Control"
author: "JoonHo Lee (jlee296@ua.edu)"
date: "`r Sys.Date()`"
description: >
  Learn to control both expected cluster count AND weight concentration 
  using the dual-anchor framework for Dirichlet Process prior elicitation.
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Beyond Cluster Counts: Dual-Anchor Elicitation for Weight Control}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
library(DPprior)
library(ggplot2)
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5,
  fig.align = "center",
  out.width = "85%"
)
set.seed(42)

# Define a consistent color palette
palette_2 <- c("#E41A1C", "#377EB8")
palette_3 <- c("#E41A1C", "#377EB8", "#4DAF4A")
palette_4 <- c("#E41A1C", "#377EB8", "#4DAF4A", "#984EA3")
```

## Overview

The previous vignettes focused on calibrating a Gamma hyperprior for $\alpha$ 
based on your expectations about the **number of clusters** $K_J$. This 
approach is intuitive: you specify how many groups you expect, and the 
DPprior package finds the corresponding prior on $\alpha$.
 
However, as [Vicentini & Jermyn (2025)](https://doi.org/10.48550/arXiv.2502.00864) 
have pointed out, matching $K_J$ alone can lead to **unintended prior 
behavior** on the cluster weights. Specifically, a prior that produces 
the "right" expected number of clusters may simultaneously imply a large
tail probability for the first size-biased cluster mass $W_{SB}$. This is not
the same estimand as the largest cluster mass $W_{max}$.

This vignette introduces the **dual-anchor framework**, which allows you 
to control both:

1. **Anchor 1**: The number of occupied clusters $K_J$ (as before)
2. **Anchor 2**: A named weight quantity (e.g., $W_{SB}$ or $\rho$) that
   captures how mass is distributed across clusters

By the end of this vignette, you will understand:

+ Why K-only calibration can produce surprising weight behavior
+ Two key weight quantities: $W_{SB}$ and $\rho$ (co-clustering probability)
+ How to use `DPprior_dual_hard()` for a verified inequality
+ How to use `DPprior_dual_soft()` for an explicit fixed-scale trade-off
+ When and how to choose between hard constraints, soft targets, and
  different $\lambda$ values


## 1. The Hidden Problem: Unintended Weight Priors

### 1.1 What K-Only Calibration Misses

Consider a researcher analyzing a multisite educational trial with 50 sites, 
expecting that treatment effects will cluster into approximately 5 distinct 
patterns. Using the K-only calibration approach from the 
[Applied Guide](applied-guide.html), they obtain:

```{r k-only-example}
J <- 50
mu_K <- 5

target_K <- DPprior_target_K(J = J, mu_K = mu_K, var_K = 8)
fit_K <- DPprior_fit(
  J = J, target_K = target_K, method = "A2-MN",
  check_diagnostics = FALSE
)
print(fit_K)
```

The fit looks reasonable: the prior expects about 5 clusters with 
appropriate uncertainty. But what does this prior imply about how 
*mass is distributed* across those clusters?

```{r k-only-weight-diagnostics}
# Examine the implied weight behavior
cat("Weight diagnostics for K-only prior:\n")
cat("  E[W_SB] =",
    round(mean_w1(fit_K$parameters$a, fit_K$parameters$b), 3), "\n")
cat("  P(W_SB > 0.3) =",
    round(prob_wsb_exceeds(0.3, fit_K$parameters$a, fit_K$parameters$b), 3), "\n")
cat("  P(W_SB > 0.5) =",
    round(prob_wsb_exceeds(0.5, fit_K$parameters$a, fit_K$parameters$b), 3), "\n")
cat("  P(W_SB > 0.9) =",
    round(prob_wsb_exceeds(0.9, fit_K$parameters$a, fit_K$parameters$b), 3), "\n")
```

Here is the key insight: even though we expect only 5 clusters on average, 
there is a **substantial probability** (nearly 50%!) that a randomly 
selected unit belongs to a cluster containing more than half of all units. 
This is much more concentrated than many researchers would intuitively 
expect from "5 clusters."

### 1.2 Why Does This Happen?

The relationship between $K_J$ and the cluster weights is not as tight 
as one might expect. The expected number of clusters follows approximately:

$$
\mathbb{E}[K_J \mid \alpha] \approx 1 + \alpha \log J
$$

while the expected first stick-breaking weight is:

$$
\mathbb{E}[w_1 \mid \alpha] = \frac{1}{1 + \alpha}
$$
 
For $J = 50$ and an expected cluster count around 5, we need 
$\alpha \approx 1$, which implies $\mathbb{E}[w_1 \mid \alpha] \approx 0.5$. 
But because $\alpha$ is *random* under our Gamma hyperprior, the tails 
of the $\alpha$ distribution can induce extreme weight behavior: small 
$\alpha$ values lead to $w_1$ close to 1, while large $\alpha$ values 
fragment the weights.

```{r mechanism-illustration, fig.height=5, fig.cap="When α is small, w₁ concentrates near 1; when α is large, w₁ concentrates near 0. A diffuse prior on α spans both extremes."}
# Illustrate the conditional relationship
alpha_grid <- c(0.5, 1, 2, 5, 10, 20)
w_grid <- seq(0.01, 0.99, length.out = 200)

cond_df <- do.call(rbind, lapply(alpha_grid, function(a) {
  data.frame(
    w = w_grid,
    density = dbeta(w_grid, 1, a),
    alpha = paste0("α = ", a)
  )
}))
cond_df$alpha <- factor(cond_df$alpha, levels = paste0("α = ", alpha_grid))

ggplot(cond_df, aes(x = w, y = density, color = alpha)) +
  geom_line(linewidth = 0.8) +
  scale_color_viridis_d(option = "plasma", end = 0.85) +
  labs(
    x = expression(w[1]),
    y = "Conditional Density",
    title = expression("Conditional Distribution of " * w[1] * " | " * alpha),
    color = NULL
  ) +
  theme_minimal() +
  theme(legend.position = "right")
```


## 2. Understanding Weight Distributions

Before diving into dual-anchor calibration, let us examine the two 
weight anchors available in the DPprior package.

### 2.1 Stick-Breaking Weights Refresher

Under Sethuraman's stick-breaking representation of the Dirichlet Process, 
the random mixing measure $G$ is:

$$
G = \sum_{h=1}^{\infty} w_h \, \delta_{\theta_h}
$$

where the weights are constructed via:

$$
v_h \overset{iid}{\sim} \text{Beta}(1, \alpha), \quad 
w_1 = v_1, \quad 
w_h = v_h \prod_{\ell < h} (1 - v_\ell) \quad (h \geq 2)
$$

The sequence $(w_1, w_2, \ldots)$ follows the **GEM($\alpha$) distribution** 
and is in *size-biased* order, not ranked by magnitude.

### 2.2 Anchor 2a: The First Stick-Breaking Weight ($w_1$)

The first weight $w_1$ has a particularly tractable distribution. 
Conditionally, $w_1 \mid \alpha \sim \text{Beta}(1, \alpha)$.

Under the hyperprior $\alpha \sim \text{Gamma}(a, b)$, the marginal 
distribution of $w_1$ has **fully closed-form** expressions:

**CDF:**
$$
F_{w_1}(x \mid a, b) = 1 - \left( \frac{b}{b - \log(1-x)} \right)^a
$$

**Quantile function:**
$$
Q_{w_1}(u \mid a, b) = 1 - \exp\left( b \left[ 1 - (1-u)^{-1/a} \right] \right)
$$

**First-size-biased-weight tail probability:**
$$
P(w_1 > t \mid a, b) = \left( \frac{b}{b - \log(1-t)} \right)^a
$$

**Important interpretability caveat:** $w_1$ is in GEM (size-biased) 
order, *not* the largest cluster weight. A faithful interpretation is:

> "$w_1$ is the asymptotic proportion of the cluster containing a 
> randomly selected unit."

This is a meaningful named tail diagnostic: if $P(W_{SB}>0.5)$ is high, then
a randomly selected unit is likely to belong to a cluster that contains more
than half the population. It is not $P(W_{max}>0.5)$.

```{r w1-functions-demo}
# Using the closed-form w1 functions
a <- fit_K$parameters$a
b <- fit_K$parameters$b

cat("W_SB distribution under K-only prior:\n")
cat("  Mean:     ", round(mean_w1(a, b), 4), "\n")
cat("  Variance: ", round(var_w1(a, b), 4), "\n")
cat("  Median:   ", round(quantile_w1(0.5, a, b), 4), "\n")
cat("  90th %ile:", round(quantile_w1(0.9, a, b), 4), "\n")
cat("\nNamed W_SB tail probabilities:\n")
cat("  P(W_SB > 0.3):", round(prob_wsb_exceeds(0.3, a, b), 4), "\n")
cat("  P(W_SB > 0.5):", round(prob_wsb_exceeds(0.5, a, b), 4), "\n")
cat("  P(W_SB > 0.9):", round(prob_wsb_exceeds(0.9, a, b), 4), "\n")
```

### 2.3 Anchor 2b: Co-Clustering Probability ($\rho$)

The co-clustering probability is defined as:

$$
\rho = \sum_{h=1}^{\infty} w_h^2
$$

This has a natural interpretation: $\rho$ equals the probability that 
two randomly selected units belong to the *same* cluster. This may be 
more intuitive for applied researchers to elicit.

The conditional moments are simple:

$$
\mathbb{E}[\rho \mid \alpha] = \frac{1}{1 + \alpha}, \quad
\text{Var}(\rho \mid \alpha) = \frac{2\alpha}{(1+\alpha)^2(2+\alpha)(3+\alpha)}
$$

Note that $\mathbb{E}[\rho \mid \alpha] = \mathbb{E}[w_1 \mid \alpha]$, 
so their *means* are equal, but their full distributions differ.

```{r rho-functions-demo}
# Using the rho (co-clustering) functions
cat("Co-clustering probability ρ under K-only prior:\n")
cat("  E[ρ]:     ", round(mean_rho(a, b), 4), "\n")
cat("  Var(ρ):   ", round(var_rho(a, b), 6), "\n")
cat("  SD(ρ):    ", round(sqrt(var_rho(a, b)), 4), "\n")

# Compare with w1
cat("\nNote: E[w₁] = E[ρ] =", round(mean_w1(a, b), 4), "\n")
cat("But: Var(w₁) =", round(var_w1(a, b), 4), "≠ Var(ρ) =", 
    round(var_rho(a, b), 6), "\n")
```


## 3. The Dual-Anchor Framework

### 3.1 The Core Idea

The v2 dual-anchor framework starts from a usable, verified K-only fit and
then selects one of two explicit contracts:

+ **Anchor 1:** Match your beliefs about $K_J$ (number of clusters)
+ **Anchor 2:** Match a named $W_{SB}$ estimand (tail probability, mean, or
  quantile), or use the separately certified $W_{max}$ tail upper bound

Use **hard mode** when Anchor 2 is a genuine inequality that must be
independently verified. It minimizes the fixed-input-scale K loss subject to
the named constraint,

$$
\min_{a>0,b>0} L_K(a,b)
\quad\text{subject to}\quad g_w(a,b) \leq 0.
$$

Use **soft mode** when both anchors are preferences and a trade-off is
scientifically appropriate:

$$
(a^*, b^*) = \arg\min_{a > 0, b > 0} \left[ 
  \lambda \cdot L_K(a, b) + 
  (1 - \lambda) \cdot L_w(a, b) 
\right]
$$

where $L_K$ measures the discrepancy from the K target and $L_w$ measures 
the discrepancy from the weight target.

Hard mode has no $\lambda$. A hard result is decision-ready only when its
mode-specific contract and the result quartet show an approved status with
`usable = TRUE` and `verified = TRUE`; the independent certificate is in
`constraint$satisfied`.

### 3.2 The Role of $\lambda$ in Soft Mode

The parameter $\lambda \in (0, 1]$ controls the soft trade-off between the
two anchors:

| $\lambda$ Value | Interpretation |
|-----------------|----------------|
| $\lambda = 1$ | K-only calibration (ignores weight anchor) |
| $0.5 \leq \lambda < 1$ | K remains primary, weight provides secondary control |
| $0 < \lambda < 0.5$ | Weight becomes primary (rare in practice) |

In most applications, we recommend $\lambda \in [0.5, 0.9]$, keeping 
the cluster count as the primary anchor while using the weight constraint 
to avoid unintended dominance behavior.

Soft-mode $\lambda$ is not a probability and never certifies an inequality.
Inspect `tradeoff`, `residuals`, and the result quartet before using a soft
fit downstream.

### 3.3 Fixed Input-Derived Scaling

The K and weight losses have different units. `DPprior_dual_soft()` records
fixed scales derived from the input fit and target, then uses those same
scales throughout optimization and verification. They are available in
`fit$tradeoff$scales` and `fit$computation$scaling`. New code does not choose
among legacy `loss_type` heuristics.


## 4. Using the v2 Dual-Anchor APIs

Begin with `DPprior_target_K()` and `DPprior_fit()`. Then choose hard or soft
mode from the meaning of the weight statement—not from which optimizer happens
to return a convenient answer.

### 4.1 Basic Usage: Verified Probability Inequality

Suppose you want to ensure that the probability of extreme dominance 
is limited. Specifically, you want $P(w_1 > 0.5) \leq 0.25$:

```{r dual-basic-prob}
# Step 1: construct and fit a canonical K target
target_K <- DPprior_target_K(J = 50, mu_K = 5, var_K = 8)
fit_K <- DPprior_fit(
  J = 50, target_K = target_K, method = "A2-MN",
  check_diagnostics = FALSE
)

cat("K-only prior:\n")
cat("  Gamma(", round(fit_K$parameters$a, 3), ",",
    round(fit_K$parameters$b, 3), ")\n")
cat("  P(w₁ > 0.5) =",
    round(prob_wsb_exceeds(
      0.5, fit_K$parameters$a, fit_K$parameters$b
    ), 3), "\n\n")

# Step 2: impose and independently verify an at-most constraint
fit_dual <- DPprior_dual_hard(
  fit = fit_K,
  constraint = list(
    metric = "wsb_tail", threshold = 0.5,
    relation = "<=", bound = 0.25
  )
)

cat("Hard dual-anchor prior:\n")
cat("  Gamma(", round(fit_dual$parameters$a, 3), ",",
    round(fit_dual$parameters$b, 3), ")\n")
cat("  P(w₁ > 0.5) =", round(fit_dual$achieved$weight$value, 3), "\n")
cat("  Constraint verified:", fit_dual$constraint$satisfied, "\n")
fit_dual[c("status", "usable", "verified", "message")]
```

Let us visualize the effect on both the $K_J$ and $w_1$ distributions:

```{r dual-comparison-viz, fig.height=6, fig.cap="Comparison of K-only vs dual-anchor priors: the dual-anchor approach shifts mass away from extreme w₁ values while maintaining similar K expectations."}
# Prepare data for visualization
logS <- compute_log_stirling(J)

# K distributions
k_grid <- 1:20
k_df <- data.frame(
  K = rep(k_grid, 2),
  probability = c(
    pmf_K_marginal(
      J, fit_K$parameters$a, fit_K$parameters$b, logS = logS
    )[k_grid],
    pmf_K_marginal(
      J, fit_dual$parameters$a, fit_dual$parameters$b, logS = logS
    )[k_grid]
  ),
  Prior = rep(c("K-only", "Dual-anchor"), each = length(k_grid))
)

# w1 distributions
w_grid <- seq(0.01, 0.95, length.out = 200)
w_df <- data.frame(
  w1 = rep(w_grid, 2),
  density = c(
    density_w1(w_grid, fit_K$parameters$a, fit_K$parameters$b),
    density_w1(w_grid, fit_dual$parameters$a, fit_dual$parameters$b)
  ),
  Prior = rep(c("K-only", "Dual-anchor"), each = length(w_grid))
)

# Create plots
p1 <- ggplot(k_df, aes(x = K, y = probability, color = Prior)) +
  geom_point(size = 1.5) +
  geom_line(linewidth = 0.8) +
  scale_color_manual(values = palette_2) +
  labs(x = "Number of Clusters (K)", y = "Probability",
       title = "Marginal PMF of K") +
  theme_minimal() +
  theme(legend.position = "bottom")

p2 <- ggplot(w_df, aes(x = w1, y = density, color = Prior)) +
  geom_line(linewidth = 1) +
  geom_vline(xintercept = 0.5, linetype = "dashed", color = "gray50") +
  scale_color_manual(values = palette_2) +
  labs(x = expression(w[1]), y = "Density",
       title = expression("Marginal Density of " * w[1])) +
  theme_minimal() +
  theme(legend.position = "bottom")

gridExtra::grid.arrange(p1, p2, ncol = 2)
```

### 4.2 Comparing Hard and Soft Semantics

The same numeric value can express different scientific claims. Here the hard
fit certifies an upper bound, while the soft fit treats 0.25 as an equality
preference with $\lambda=0.5$:

```{r hard-soft-comparison}
soft_target <- list(
  metric = "wsb_tail", threshold = 0.5,
  relation = "target", value = 0.25
)
fit_soft <- DPprior_dual_soft(
  fit = fit_K, target = soft_target, lambda = 0.5
)

fits_compare <- list(K_only = fit_K, Hard = fit_dual, Soft = fit_soft)
hard_soft_compare <- data.frame(
  Method = names(fits_compare),
  a = vapply(fits_compare, function(x) x$parameters$a, numeric(1)),
  b = vapply(fits_compare, function(x) x$parameters$b, numeric(1)),
  E_K = vapply(fits_compare, function(x) x$achieved$K$mean, numeric(1)),
  P_w1_gt_50 = vapply(fits_compare, function(x) {
    prob_wsb_exceeds(0.5, x$parameters$a, x$parameters$b)
  }, numeric(1)),
  Status = vapply(fits_compare, function(x) x$status, character(1)),
  Usable = vapply(fits_compare, function(x) x$usable, logical(1)),
  Verified = vapply(fits_compare, function(x) x$verified, logical(1))
)

knitr::kable(
  hard_soft_compare,
  digits = 3,
  caption = paste(
    "Hard inequality versus fixed-scale soft target for",
    "P(w₁ > 0.5) = 0.25"
  )
)
```

Only the hard result has `constraint$satisfied`; a soft result instead records
component and total losses in `tradeoff`.

### 4.3 Alternative Hard Constraint: Quantile

You can also specify a quantile constraint. For example, to ensure that 
the 90th percentile of $w_1$ does not exceed 0.6:

```{r dual-quantile}
fit_dual_q <- DPprior_dual_hard(
  fit = fit_K,
  constraint = list(
    metric = "wsb_quantile", probability = 0.9,
    relation = "<=", bound = 0.6
  )
)

cat("Hard dual-anchor (quantile constraint):\n")
cat("  Gamma(", round(fit_dual_q$parameters$a, 3), ",",
    round(fit_dual_q$parameters$b, 3), ")\n")
cat("  90th percentile of w₁:",
    round(fit_dual_q$achieved$weight$value, 3), "\n")
cat("  Constraint verified:", fit_dual_q$constraint$satisfied, "\n")
```

### 4.4 Alternative Soft Target: Mean

For constraints based on $\mathbb{E}[w_1]$:

```{r dual-mean}
fit_dual_m <- DPprior_dual_soft(
  fit = fit_K,
  target = list(
    metric = "wsb_mean", relation = "target", value = 0.35
  ),
  lambda = 0.6
)

cat("Soft dual-anchor (mean target):\n")
cat("  Gamma(", round(fit_dual_m$parameters$a, 3), ",",
    round(fit_dual_m$parameters$b, 3), ")\n")
cat("  E[w₁]:", round(fit_dual_m$achieved$weight$value, 3), "\n")
cat("  (Target was 0.35)\n")
print(fit_dual_m$tradeoff[c("lambda", "K_loss", "weight_loss", "total_loss")])
fit_dual_m[c("status", "usable", "verified", "message")]
```


## 5. Elicitation Questions for Weight Anchors

Translating substantive knowledge into weight constraints requires 
careful framing. Here are suggested elicitation questions:

### For $w_1$ (Size-Biased Weight)

These questions focus on what happens when you randomly sample a unit:

> "If you randomly select a site from your study, what proportion of 
> all sites would you expect to share its effect pattern? What's a 
> reasonable median estimate?"

> "Is there a significant chance (say, > 30%) that a randomly selected 
> site belongs to an effect group that contains more than half of all sites?"

### For $\rho$ (Co-Clustering Probability)

These questions may be more intuitive for many researchers:
 
> "If you pick two sites at random, what's the probability that they 
> belong to the same effect group?"

> "How likely is it that any two randomly chosen sites would show 
> substantively similar treatment effects?"

### Translation Examples

| Elicited Belief | Formal Constraint |
|-----------------|-------------------|
| "Randomly selected site's group is at most 25% (median)" | Hard: `metric="wsb_quantile", probability=.5, relation="<=", bound=.25` |
| "At most 20% chance that a random site is in a dominant (>50%) group" | Hard: `metric="wsb_tail", threshold=.5, relation="<=", bound=.2` |
| "Two random sites have about 20% chance of the same group" | Diagnose $\rho$ separately; do not relabel it as a v2 $W_{SB}$ target |


## 6. Exploring the Trade-off: $\lambda$ Sensitivity

Different values of $\lambda$ produce different trade-offs between 
matching the K anchor and the weight anchor. Let us explore this 
systematically:

```{r lambda-sensitivity, fig.height=5, fig.cap="As λ decreases (giving more weight to the w₁ constraint), P(w₁ > 0.5) decreases but the K distribution may shift."}
lambda_grid <- c(1.0, 0.9, 0.7, 0.5, 0.3)
w1_target <- list(
  metric = "wsb_tail", threshold = 0.5,
  relation = "target", value = 0.2
)

lambda_results <- lapply(lambda_grid, function(lam) {
  fit <- DPprior_dual_soft(
    fit = fit_K, target = w1_target, lambda = lam
  )
  
  list(
    lambda = lam,
    a = fit$parameters$a,
    b = fit$parameters$b,
    E_K = fit$achieved$K$mean,
    P_w1_gt_50 = prob_wsb_exceeds(
      0.5, fit$parameters$a, fit$parameters$b
    ),
    E_w1 = mean_w1(fit$parameters$a, fit$parameters$b),
    status = fit$status,
    usable = fit$usable,
    verified = fit$verified
  )
})

# Create comparison table
lambda_df <- do.call(rbind, lapply(lambda_results, function(r) {
  data.frame(
    lambda = r$lambda,
    a = round(r$a, 3),
    b = round(r$b, 3),
    E_K = round(r$E_K, 2),
    P_w1_gt_50 = round(r$P_w1_gt_50, 3),
    E_w1 = round(r$E_w1, 3),
    status = r$status,
    usable = r$usable,
    verified = r$verified
  )
}))

knitr::kable(
  lambda_df,
  col.names = c(
    "λ", "a", "b", "E[K]", "P(w₁>0.5)", "E[w₁]",
    "Status", "Usable", "Verified"
  ),
  caption = paste(
    "Effect of λ under the fixed-scale soft contract",
    "(target: P(w₁>0.5) = 0.2)"
  )
)
```

```{r lambda-tradeoff-viz, fig.height=4, fig.cap="Trade-off curve: reducing λ brings P(w₁ > 0.5) closer to target at the cost of K deviation."}
# Visualize the trade-off
tradeoff_df <- data.frame(
  lambda = sapply(lambda_results, `[[`, "lambda"),
  E_K = sapply(lambda_results, `[[`, "E_K"),
  P_w1 = sapply(lambda_results, `[[`, "P_w1_gt_50")
)

ggplot(tradeoff_df, aes(x = E_K, y = P_w1)) +
  geom_path(color = "gray50", linewidth = 0.8) +
  geom_point(aes(color = factor(lambda)), size = 4) +
  geom_hline(yintercept = 0.2, linetype = "dashed", color = "#E41A1C", alpha = 0.7) +
  geom_vline(xintercept = 5, linetype = "dashed", color = "#377EB8", alpha = 0.7) +
  annotate("text", x = 4.7, y = 0.22, label = "w₁ target", hjust = 1, color = "#E41A1C") +
  annotate("text", x = 5.1, y = 0.45, label = "K target", hjust = 0, color = "#377EB8") +
  scale_color_viridis_d(option = "viridis", begin = 0.2, end = 0.8) +
  labs(
    x = "E[K] (Cluster Count)",
    y = expression(P(w[1] > 0.5)),
    title = "Trade-off Between K and Weight Anchors",
    color = "λ"
  ) +
  theme_minimal() +
  theme(legend.position = "right")
```


## 7. Practical Recommendations

### 7.1 When to Use Dual-Anchor Calibration

Consider dual-anchor calibration when:

+ **A predeclared $W_{SB}$ policy is triggered:** For example, the reviewed
  fit has $P(W_{SB}>0.5)$ above an application-specific action threshold
  
+ **Strong prior belief against concentration:** You believe effects 
  should be reasonably spread across groups, not concentrated
  
+ **Low-information settings:** When the data may not strongly update 
  the prior, unintended weight behavior can persist to the posterior

Do **not** use dual-anchor when:

+ You are comfortable with the weight implications of your K-only prior
+ You have strong data that will overwhelm any prior specification
+ The additional complexity is not justified for your application

### 7.2 Choosing Hard or Soft Mode

| Scientific statement | API | Evidence to inspect |
|----------------------|-----|---------------------|
| A weight inequality must hold | `DPprior_dual_hard()` | `constraint$satisfied`, residual, tolerance, verification |
| K and weight targets are competing preferences | `DPprior_dual_soft()` | `tradeoff`, residuals, result quartet |

Do not report a soft target as if it were a hard constraint. Conversely, do
not introduce $\lambda$ into a hard inequality.

### 7.3 Choosing $\lambda$ in Soft Mode

| Situation | Recommended $\lambda$ |
|-----------|----------------------|
| K anchor is well-justified, weight is secondary constraint | 0.8 – 0.9 |
| Both anchors are equally important | 0.5 – 0.7 |
| Weight behavior is primary concern | 0.3 – 0.5 |

In practice, we recommend starting with $\lambda = 0.7$ and adjusting 
based on how closely each target is achieved.

### 7.4 Choosing a Supported Weight Estimand

| Estimand | Advantages | Best For |
|----------|------------|----------|
| $W_{SB}$ tail | Closed form; explicitly size-biased | Probability inequalities or targets |
| $W_{SB}$ mean | Simple scalar interpretation | Mean preference or bound |
| $W_{SB}$ quantile | Closed form | Quantile preference or bound |
| certified $W_{max}$ tail upper bound | Conservative certified evidence | At-most safety constraints |

The co-clustering quantity $\rho$ remains a useful diagnostic, and
$\mathbb{E}[w_1]=\mathbb{E}[\rho]$, but the v2 hard/soft public target grammar
does not silently relabel a $W_{SB}$ claim as a direct $\rho$ claim.


## 8. When Dual-Anchor Constraints Are Infeasible

Not all combinations of K and weight constraints are achievable. Hard mode
does not return a failed inequality as if it were satisfied. It raises a typed
condition whose `result` retains the complete canonical evidence:

```{r infeasible-example}
# An aggressive weight target that conflicts with K target
hard_condition <- NULL
fit_aggressive <- tryCatch(
  DPprior_dual_hard(
    fit = fit_K,
    constraint = list(
      metric = "wsb_tail", threshold = 0.5,
      relation = "<=", bound = 0.05
    )
  ),
  error = function(condition) {
    hard_condition <<- condition
    condition$result
  }
)

cat("Aggressive dual-anchor attempt:\n")
cat("  Target: P(w₁ > 0.5) = 0.05\n")
fit_aggressive[c("status", "usable", "verified", "message")]
if (!is.null(fit_aggressive$parameters)) {
  cat("  Achieved: P(w₁ > 0.5) =",
      round(fit_aggressive$achieved$weight$value, 3), "\n")
  cat("  Constraint satisfied:",
      fit_aggressive$constraint$satisfied, "\n")
}
if (!is.null(hard_condition)) {
  cat("  Typed condition:", class(hard_condition)[1], "\n")
}
```

When constraints are too strict, consider:

1. **Relaxing one of the targets:** Accept a higher named $W_{SB}$ tail
   probability or a different K expectation
   
2. **Changing the scientific contract:** If compromise is defensible, use
   soft mode and perform an explicit $\lambda$ sensitivity analysis
   
3. **Using a single anchor:** Sometimes K-only or weight-only 
   calibration is more appropriate


## Summary

| Concept | Key Point |
|---------|-----------|
| K-only calibration | May induce unintended weight behavior |
| $w_1$ (first weight) | Size-biased cluster mass; closed-form distribution |
| $\rho$ (co-clustering) | P(two random units share a cluster); intuitive |
| Hard dual anchor | Verified named inequality; no $\lambda$ |
| Soft dual anchor | Fixed-scale trade-off between K and weight fit via $\lambda$ |
| Recommended workflow | 1. `DPprior_target_K()` → 2. `DPprior_fit()` → 3. hard or soft dual API |

## Legacy v2.x Compatibility

`DPprior_dual()` is retained throughout v2.x for existing equality-loss
workflows. It is an approximate legacy adapter, not a hard-constraint
certificate: its canonical result is visibly `status = "approximate"`,
`usable = TRUE`, and `verified = FALSE`. It will not be removed before v3.0
and a migration review. New analyses should use `DPprior_dual_hard()` or
`DPprior_dual_soft()`.

```{r legacy-dual-v2, eval=FALSE}
# Compatibility only; do not interpret this as a verified inequality.
legacy_fit <- DPprior_dual(
  fit = fit_K,
  w1_target = list(prob = list(threshold = 0.5, value = 0.25)),
  lambda = 0.7,
  loss_type = "adaptive"
)
legacy_fit[c("status", "usable", "verified", "message")]
```

## What's Next?

+ **[Diagnostics](diagnostics.html)**: Deep dive into verifying your 
  prior meets specifications
  
+ **[Case Studies](case-studies.html)**: Real-world examples applying 
  dual-anchor calibration
  
+ **[Mathematical Foundations](theory-overview.html)**: Theoretical 
  details of the DPprior framework


## References

Vicentini, S., & Jermyn, I. H. (2025). Prior selection for the precision 
parameter of Dirichlet process mixture models. *arXiv:2502.00864*. 
https://doi.org/10.48550/arXiv.2502.00864

Sethuraman, J. (1994). A constructive definition of Dirichlet priors. 
*Statistica Sinica*, 4(2), 639–650.

---

*For questions or feedback, please visit the 
[GitHub repository](https://github.com/joonho112/DPprior).*
