---
title: "Quick Start: Your First Prior in 5 Minutes"
author: "JoonHo Lee (jlee296@ua.edu)"
date: "`r Sys.Date()`"
description: >
  Learn how to elicit your first Gamma hyperprior for α in a Dirichlet Process 
  model in just 5 minutes using the DPprior package.
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Quick Start: Your First Prior in 5 Minutes}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
library(DPprior)
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5,
  fig.align = "center",
  out.width = "85%"
)
set.seed(42)
```

## Overview

This vignette demonstrates the fastest path to eliciting a Gamma hyperprior 
for the concentration parameter $\alpha$ in a Dirichlet Process mixture model. 
By the end of this 5-minute tutorial, you will be able to:
 
1. Specify your prior belief about the number of clusters
2. Convert that belief into Gamma hyperparameters
3. Visualize and verify your elicited prior

## 1. Minimal Example

Let's walk through a concrete scenario. Suppose you are analyzing a multisite 
educational trial with 50 sites, and you expect that these sites will cluster 
into approximately 5 distinct effect groups. This setup mirrors the "DP-inform" 
prior elicitation approach used in 
[Lee et al. (2025)](https://doi.org/10.3102/10769986241254286), who demonstrated 
the benefits of informative Dirichlet Process priors for estimating site-specific 
effects in multisite trials. While that paper relied on chi-square discrepancy 
measures with computationally intensive grid search, the DPprior package 
provides fast calibrated priors for the same elicitation task, using exact
moment matching by default with a closed-form approximation available for rapid
exploration.

```{r minimal-example}
library(DPprior)

# Just two inputs needed:
# 1. J: Number of sites (or observations for clustering)
# 2. mu_K: Expected number of clusters

fit <- DPprior_fit(J = 50, mu_K = 5, confidence = "medium")

# View the result
print(fit)
```

That's it! You now have a principled Gamma hyperprior for $\alpha$.

For a reusable or reviewable specification, construct the same canonical
cluster-count target explicitly and pass it to the wrapper:

```{r canonical-target}
target_K <- DPprior_target_K(J = 50, mu_K = 5, confidence = "medium")
fit_from_target <- DPprior_fit(
  J = 50,
  target_K = target_K,
  method = "A2-MN",
  check_diagnostics = FALSE
)
fit_from_target$target$K$implied[c("mean", "variance")]
```

## 2. Understanding the Output

The `DPprior_fit()` function returns a canonical `dpprior.result/1` object.
Its scientific fields are nested; compatibility aliases are not authoritative.
Let's examine the key components:

```{r output-components}
# The Gamma hyperparameters
cat("Gamma shape (a):", round(fit$parameters$a, 4), "\n")
cat("Gamma rate  (b):", round(fit$parameters$b, 4), "\n")

# What these imply about alpha
alpha_mean <- fit$parameters$a / fit$parameters$b
alpha_sd <- sqrt(fit$parameters$a) / fit$parameters$b
cat("\nPrior mean of α:", round(alpha_mean, 3), "\n")
cat("Prior SD of α:  ", round(alpha_sd, 3), "\n")

# The target specification and decision fields
cat("\nTarget E[K]:", fit$target$K$implied$mean, "\n")
cat("Target Var(K):", round(fit$target$K$implied$variance, 2), "\n")
fit[c("status", "usable", "verified")]
```

**Interpretation**: The elicited prior $\alpha \sim \text{Gamma}(a, b)$ implies 
that you expect around `r round(fit$target$K$implied$mean, 1)` clusters when sampling
$J = `r fit$J`$ observations, with uncertainty captured by the variance.

## 3. Visualizing Your Prior

The DPprior package provides built-in visualization functions to help you 
understand and communicate your elicited prior.

### 3.1 Full Dashboard View

The `plot()` method displays a comprehensive four-panel dashboard:

```{r fig-dashboard, fig.cap="Prior elicitation dashboard showing the distributions of α, K, and w₁.", fig.height=7}
plot(fit)
```

The dashboard includes:

- **Panel (A)**: The Gamma prior density for the concentration parameter $\alpha$, 
  with key statistics (mean, CV, 95% CI)
- **Panel (B)**: The implied marginal PMF of the number of clusters $K_J$, 
  showing E[K], Var(K), and mode
- **Panel (C)**: The prior density of the first stick-breaking weight $w_1$, 
  with dominance risk assessment (useful for detecting overly concentrated priors)
- **Panel (D)**: Summary statistics table consolidating all key diagnostics

### 3.2 Individual Plots

You can also create individual plots for more focused analysis:

```{r fig-alpha1, fig.cap="Prior density for the concentration parameter α.", eval=FALSE}
plot_alpha_prior(fit)
```

```{r fig-alpha2, echo=FALSE, fig.cap="Prior density for the concentration parameter alpha.", fig.alt="A smooth right-skewed density curve showing the Gamma prior on the concentration parameter alpha."}
invisible(plot_alpha_prior(fit))
```



```{r fig-K1, fig.cap="Marginal PMF of the number of clusters K.", eval =FALSE}
plot_K_prior(fit)
```

```{r fig-K2, echo=FALSE, fig.cap="Marginal PMF of the number of clusters K.", fig.alt="A bar plot of the prior probability mass for the number of clusters, centered near the target mean."}
invisible(plot_K_prior(fit))
```


```{r fig-w1_!, fig.cap="Distribution of the first stick-breaking weight w₁.", eval=FALSE}
plot_w1_prior(fit)
```

```{r fig-w1_2, echo=FALSE, fig.cap="Distribution of the first stick-breaking weight w1.", fig.alt="A density plot for the first stick-breaking weight with visual reference lines for dominance thresholds."}
invisible(plot_w1_prior(fit))
```

## 4. Three Ways to Specify Your Uncertainty

The `DPprior_fit()` function offers flexibility in how you express your prior 
uncertainty about the number of clusters:

### Method 1: Confidence Level (Recommended for Beginners)

The simplest approach uses qualitative confidence levels. The low-confidence
case is especially diffuse, so the example requests 160 quadrature nodes and
requires the package's independent higher-order diagnostic check to pass.

```{r method-confidence}
# Low confidence = high uncertainty (wide prior). Use a higher quadrature
# order so this diffuse prior's diagnostics pass independent verification.
fit_low <- DPprior_fit(
  J = 50, mu_K = 5, confidence = "low", M = 160L
)

# Medium confidence = moderate uncertainty (default)
fit_med <- DPprior_fit(J = 50, mu_K = 5, confidence = "medium")

# High confidence = low uncertainty (concentrated prior)
fit_high <- DPprior_fit(J = 50, mu_K = 5, confidence = "high")

# Compare the resulting Gamma parameters
comparison <- data.frame(
  Confidence = c("Low", "Medium", "High"),
  var_K = round(c(fit_low$target$K$implied$variance,
                  fit_med$target$K$implied$variance,
                  fit_high$target$K$implied$variance), 2),
  a = round(c(fit_low$parameters$a, fit_med$parameters$a,
              fit_high$parameters$a), 3),
  b = round(c(fit_low$parameters$b, fit_med$parameters$b,
              fit_high$parameters$b), 3),
  CV_alpha = round(1/sqrt(c(fit_low$parameters$a,
                            fit_med$parameters$a,
                            fit_high$parameters$a)), 3)
)
comparison
```

**Interpretation**: Higher confidence (lower uncertainty) leads to a more 
concentrated prior on $\alpha$, reflected in the lower coefficient of 
variation (CV).

### Method 2: Direct Variance Specification

For users who want precise control:

```{r method-variance}
# Specify variance of K directly
fit_direct <- DPprior_fit(J = 50, mu_K = 5, var_K = 10)
cat("Direct specification: var_K = 10\n")
cat("  Gamma(a =", round(fit_direct$parameters$a, 3),
    ", b =", round(fit_direct$parameters$b, 3), ")\n")
```

### Method 3: Different Calibration Methods

For advanced users, different calibration algorithms are available:

```{r method-algorithms}
# A2-MN (default): Exact moment matching via Newton's method
fit_newton <- DPprior_fit(J = 50, mu_K = 5, var_K = 10, method = "A2-MN")

# A1: Fast closed-form approximation (good for large J)
fit_approx <- DPprior_fit(J = 50, mu_K = 5, var_K = 10, method = "A1")

cat("A2-MN: Gamma(", round(fit_newton$parameters$a, 3), ", ",
    round(fit_newton$parameters$b, 3), ")\n", sep = "")
cat("A1:    Gamma(", round(fit_approx$parameters$a, 3), ", ",
    round(fit_approx$parameters$b, 3), ")\n", sep = "")
```

## 5. Using Your Prior in Practice

Once you have elicited your Gamma hyperparameters, you can use them in your 
Bayesian software of choice.

### Stan
```stan
data {
  int<lower=1> J;
}
parameters {
  real<lower=0> alpha;
}
model {
  // Elicited prior from DPprior
  alpha ~ gamma(1.6, 0.816);  // Use fit$parameters$a and $b from R
}
```

### JAGS
```r
model {
  # Elicited prior from DPprior
  alpha ~ dgamma(1.6, 0.816)  # Note: JAGS uses rate parameterization
  
  # ... rest of your DP mixture model
}
```

### R (for simulation)
```{r r-usage}
# Draw samples from the elicited prior
n_samples <- 10000
alpha_samples <- rgamma(
  n_samples,
  shape = fit$parameters$a,
  rate = fit$parameters$b
)

cat("Summary of sampled α values:\n")
cat("  Mean:", round(mean(alpha_samples), 3), "\n")
cat("  SD:  ", round(sd(alpha_samples), 3), "\n")
cat("  95% CI: [", round(quantile(alpha_samples, 0.025), 3), ", ",
    round(quantile(alpha_samples, 0.975), 3), "]\n", sep = "")
```

## What's Next?

This quick start covered the essentials. For more advanced topics:

- Use `DPprior_dual_hard()` for a verified weight inequality and
  `DPprior_dual_soft()` for an explicit fixed-scale trade-off. The retained
  `DPprior_dual()` function is a legacy equality-loss adapter, not a hard
  constraint certificate; it remains available throughout v2.x and will not
  be removed before v3.0 and a migration review.
 
- **[Applied Guide](applied-guide.html)**: Comprehensive elicitation workflow 
  with sensitivity analysis
- **[Dual-Anchor Framework](dual-anchor.html)**: Control both cluster count AND 
  cluster weight concentration
- **[Diagnostics](diagnostics.html)**: Verify your prior meets your specifications
- **[Case Studies](case-studies.html)**: Real-world examples from education, 
  medicine, and policy research

## Summary

| Step | Action | Code |
|------|--------|------|
| 1 | Define context | `J <- 50` |
| 2 | Set expected clusters | `mu_K <- 5` |
| 3 | Choose confidence | `confidence = "medium"` |
| 4 | Elicit prior | `fit <- DPprior_fit(J = J, mu_K = mu_K, confidence = "medium")` |
| 5 | Visualize | `plot(fit)` |
| 6 | Use in model | `alpha ~ gamma(fit$parameters$a, fit$parameters$b)` |

## References

Lee, J., Che, J., Rabe-Hesketh, S., Feller, A., & Miratrix, L. (2025). 
Improving the estimation of site-specific effects and their distribution in 
multisite trials. *Journal of Educational and Behavioral Statistics*, 50(5), 
731–764. https://doi.org/10.3102/10769986241254286



---

*For questions or feedback, please visit the [GitHub repository](https://github.com/joonho112/DPprior).*
